---
title: "Project Econometrics"
author: "Emile Guillaume, Jean Lavallée and Aurélian Moge"
date: "2026-04-15"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      cache = TRUE, fig.width = 7, fig.height = 4)
```

# Packages, figure parameters, functions and data

```{r, include=FALSE}
pkgs <- c("haven", "tidyverse", "sjlabelled", "modelsummary",
          "AER", "sandwich", "lmtest", "car", "plm", "Formula", "broom",
          "gridExtra", "margins", "boot")
for (p in pkgs) if (!requireNamespace(p, quietly = TRUE)) install.packages(p)
library(haven); library(tidyverse); library(sjlabelled)
library(modelsummary); library(AER); library(sandwich)
library(lmtest); library(car); library(Formula); library(broom)
library(gridExtra); library(plm); library(margins); library(boot)

# Function for figures theme
theme_projet <- function() {
  theme_bw() +
    theme(
      panel.border       = element_blank(),
      axis.line          = element_line(color = "black"),
      panel.grid.major.x = element_blank(),
      panel.grid.minor   = element_blank()
    )
}

# Function for household-clustered robust (HC1) variance
vcov_robust <- function(m) {
  vcovCL(m, cluster = ~ id_menage, type = "HC1")
}

# Function for robust Wald test that all origin dummies are jointly zero
wald_test_origin <- function(m) {
  nms <- grep("^origine", names(coef(m)), value = TRUE)
  if (length(nms) == 0) return(NA_real_)
  test <- linearHypothesis(m, nms, vcov. = vcov_robust(m),
                           # Under a robust vcov we use the Chi-square
                           test = "Chisq")
  test[2, "Pr(>Chisq)"]
}

# Function to summarize gamma hat
summarise_gamma <- function(mod, name) {
  co <- coef(summary(mod))
  row <- grep("lag", rownames(co))[1]
  tibble(estimator = name,
         gamma_hat = co[row, 1],
         se        = co[row, 2])
}
```

```{r}
df <- read_dta("DM_Subject_2_Data.dta") %>%
  mutate(
    # Factor variables
    origine        = factor(origine),
    immi           = factor(immi),
    descimmi       = factor(descimmi),
    homme          = factor(homme),
    education      = factor(education),
    csp_actif      = factor(csp_actif),
    matri          = factor(matri),
    zus            = factor(zus),
    tuu_r          = factor(tuu_r),
    region         = factor(region),
    sante_declaree = factor(sante_declaree),
    contrat        = factor(contrat),
    typehoraire    = factor(typehoraire),
    compenheuresup = factor(compenheuresup),
    csp_pere_gp    = factor(substr(as.character(csp_pere), 1, 1)),
    csp_mere_gp    = factor(substr(as.character(csp_mere), 1, 1)),
    # Correct the "salhoraire" variable
    salhoraire     = salmensuel/(heuretra * 4.33),
    # Replace numeric codes with readable labels
    origine        = fct_recode(origine,
                       "France"                  = "01",
                       "Northern Europe"         = "03",
                       "Southern Europe"         = "04",
                       "Eastern Europe"          = "05",
                       "Maghreb"                 = "06",
                       "Rest of Africa"          = "07",
                       "Middle East"             = "08",
                       "Southeast Asia"          = "09",
                       "RoW"                     = "10",
                       "Unknown"                 = "99"
                     ),
    # France as reference
    origine        = relevel(origine, ref = "France")
  )
```

# Question 1: Cross-section of wages for employed individuals

## Sample

In the general case, the wage is registered for the first and the last
quarters ($T=1,2$), among six quarters ($t=1,2,3,4,5,6$). Here, we
restrict our sample to employed individuals (*i.e.*, `acteu == "1"`) in
the last quarter (*i.e.*, $T=2$ or $t=6$), for whom hourly wage is
available (*i.e.*, `!is.na(salhoraire)`).

```{r}
df_q1 <- df %>% filter(t==6, acteu == "1", !is.na(salhoraire))
```

We also inspect individuals interviewed only twice (`rim` indicator, cf.
hint in the appendix).

```{r}
with(df_q1, table(rim, cut(age, c(-Inf, 30, 50, 65, Inf))))
```

Eventually, we restrict our sample to observations where no covariate is
missing

```{r}
covariates <- c("salhoraire", "logsalhoraire",
                 "origine", "immi", "age", "homme", "education",
                 "region", "tuu_r", "zus", "matri", "sante_declaree",
                 "csp_actif", "contrat", "typehoraire",
                 "csp_pere_gp", "csp_mere_gp",
                 "expe_specifique", "id_menage")
mask <- complete.cases(df_q1[, covariates])
df_q1 <- df_q1[mask, ]
nrow(df_q1)
```

After filtering observations, we run some descriptive statistics to get
a good understanding of the dataset.

```{r}
# Origin
df_q1 %>%
  count(origine) %>%
  mutate(origine = fct_reorder(origine, n)) %>%
  ggplot(aes(x = origine, y = n)) +
  geom_col(fill = "#248BFF") +
  geom_text(aes(label = n), hjust = -0.15, size = 3.2) +
  coord_flip(clip = "off") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Distribution of employed individuals by origin (t=6)",
       x = "Origin", y = "Count") +
  theme_projet()

# Wages
df_q1 %>%
  ggplot(aes(x = logsalhoraire)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50,
                 fill = "#248BFF", colour = "white") +
  geom_density(colour = "#e81757", linewidth = 0.8) +
  labs(title = "Distribution of log hourly wages (t=6)",
       x = "Log hourly wage", y = "Density") +
  theme_projet()

# Log wages by origin
df_q1 %>%
  ggplot(aes(x = origine, y = logsalhoraire)) +
  geom_boxplot(fill = "#248BFF", colour = "#248BFF", alpha=0.2, outlier.size = 0.8) +
  labs(title = "Log hourly wages by origin (t=6)",
       x = "Origin", y = "Log hourly wage") +
  theme_projet()
```

The descriptive statistics reveal two key features of the data. First,
the origin distribution is skewed: the vast majority of employed
individuals at $t=6$ are of French origin, with relatively small
subgroups for each foreign origin. This imbalance will affect the
precision of the origin coefficients. Second, the distribution of log
hourly wages is approximately normal, supporting the use of a log-level
specification. The boxplots show moderate variation in median wages
across origin groups, but with substantial overlap, suggesting that any
origin gap must be disentangled from other determinants.

## 1a) Estimation of the effect of some variables on wages

### First results

To begin with, we try a few specifications were we regress `salhoraire`
and `logsalhoraire` on variables that are commonly used in the
literature to explain wages, i.e., sequentially: age, gender, education,
location and socio-professional category. We also include `origin` in
all models as it is the independent variable we are interested in.

We first run specifications in levels

```{r}
# (1) Origin only
reg1_lvl <- lm(salhoraire ~ origine, df_q1)

# (2) Mincer-like specification = (1) + Mincer covariates
reg2_lvl <- lm(salhoraire ~ origine + age + I(age^2) + homme + education + region, df_q1)

# (3) = (2) + individual and family backgrounds
reg3_lvl <- lm(salhoraire ~ origine + age + I(age^2) + homme + education + region + expe_specifique + csp_pere_gp + csp_mere_gp + matri + sante_declaree + tuu_r + zus, df_q1)

# (4) = (3) + job characteristics
reg4_lvl <- lm(salhoraire ~ origine + age + I(age^2) + homme + education + region + expe_specifique + csp_pere_gp + csp_mere_gp + matri + sante_declaree + tuu_r + zus + csp_actif + contrat + typehoraire, df_q1)
```

And then use specifications in log

```{r}
# (5) Origin only
reg5_log <- lm(logsalhoraire ~ origine, df_q1)

# (6) Mincer-like specification = (5) + Mincer covariates
reg6_log <- lm(logsalhoraire ~ origine + age + I(age^2) + homme + education + region, df_q1)

# (7) = (6) + individual and family backgrounds
reg7_log <- lm(logsalhoraire ~ origine + age + I(age^2) + homme + education + region + expe_specifique + csp_pere_gp + csp_mere_gp + matri + sante_declaree + tuu_r + zus, df_q1)

# (8) = (7) + job characteristics
reg8_log <- lm(logsalhoraire ~ origine + age + I(age^2) + homme + education + region + expe_specifique + csp_pere_gp + csp_mere_gp + matri + sante_declaree + tuu_r + zus + csp_actif + contrat + typehoraire, df_q1)
```

We display the results in a single table

```{r}
models <- list(
  "(1)" = reg1_lvl,
  "(2)" = reg2_lvl,
  "(3)" = reg3_lvl,
  "(4)" = reg4_lvl,
  "(5)" = reg5_log,
  "(6)" = reg6_log,
  "(7)" = reg7_log,
  "(8)" = reg8_log
  )

rows <- tribble(
  ~term,              ~"(1)", ~"(2)", ~"(3)", ~"(4)", ~"(5)", ~"(6)", ~"(7)", ~"(8)",
  "Education FE",     "",   "Yes",  "Yes",  "Yes",  "",   "Yes",  "Yes",  "Yes",
  "Region FE",        "",   "Yes",  "Yes",  "Yes",  "",   "Yes",  "Yes",  "Yes",
  "Matri FE",         "",   "",   "Yes",  "Yes",  "",   "",   "Yes",  "Yes",
  "Santé déclarée FE","",   "",   "Yes",  "Yes",  "",   "",   "Yes",  "Yes",
  "TUU_r FE",         "",   "",   "Yes",  "Yes",  "",   "",   "Yes",  "Yes",
  "ZUS FE",           "",   "",   "Yes",  "Yes",  "",   "",   "Yes",  "Yes",
  "CSP père FE",      "",   "",   "Yes",  "Yes",  "",   "",   "Yes",  "Yes",
  "CSP mère FE",      "",   "",   "Yes",  "Yes",  "",   "",   "Yes",  "Yes",
  "CSP actif FE",     "",   "",   "",   "Yes",  "",   "",   "",   "Yes",
  "Contrat FE",       "",   "",   "",   "Yes",  "",   "",   "",   "Yes",
  "Type horaire FE",  "",   "",   "",   "Yes",  "",   "",   "",   "Yes"
)

modelsummary(
  models,
  vcov = vcov_robust,
  stars = TRUE,
  output = "markdown",
  coef_omit = "^\\(Intercept\\)|^education|^region|^matri|^sante_declaree|^tuu_r|^zus|^csp_pere_gp|^csp_mere_gp|^csp_actif|^contrat|^typehoraire",
  title = NULL,
  gof_omit = "IC|Log|F|RMSE",
  add_rows = rows
  )
```

In all specifications, we report robust standard errors clustered at the
household level (`id_menage`).

We obtain models where some of the coefficients are statistically
significant at the 10%, 5% and 1% thresholds, and where the adjusted
R-squared falls between 0.004 (specification (1)) and 0.410
(specification (8)). We detail below our justification for each included
variable.

We also run a joint Wald test on origin dummies. Individual origin
coefficients are often non-significant due to small subgroup sizes, but
the joint test tells us whether origin as a whole is a statistically
meaningful determinant of wages. The joint Wald test checks whether all
origin dummies are simultaneously zero. This is more informative than
looking at individual *p*-values, especially when subgroups are small as
in this setting.

```{r}
wald_pvals <- tibble(
  Specification = names(models),
  `Wald p-value (origin)` = sapply(models, wald_test_origin)
)
knitr::kable(wald_pvals, digits = 3)
```

The Wald test of joint nullity of all origin dummies rejects in
specifications (1) and (5) (*p*-val. \< 0.01). As controls are added,
the *p*-value increases but typically remains below or near 0.05 (except
for specifications (3) and (4)), indicating that origin as a whole
remains a statistically meaningful determinant of wages even after
controlling for the usual covariates.

### Justification of control variables

Our goal is to obtain an unbiased estimator of the effect of `origine`
on wages. This requires $E[\varepsilon|\text{origine}] = 0$, meaning
there must be no omitted variables correlated with both origin and
wages. We therefore include controls that are plausibly correlated with
both.

One thing we want when estimating the effect of origin on individual's
wages by linear regression is to find an unbiased estimator of the
parameter associated to origin in such a regression. For that to be
verified, we need:

1.  The linearity in the parameter (we will specify our models to do
    so).

2.  The sample to be representative of the population (we have no hold
    over it but we may discuss it later).

3.  The non-perfect colinearity of the independent variables we are
    going to choose in our model (we will have to check it for each
    added variable).

4.  The nullity of the conditional expectation of the error term of the
    regression ($E[\varepsilon|origin]=0$).

Our two minimal theoretical models are
$\log(salhoraire) = \beta_0 + \beta_1 origin + \varepsilon$ (column (1))
and $salhoraire = \delta_0 + \delta_1 origin + \varepsilon$ (column
(5)). Our alternative specifications progressively add:

-   **(2)/(6) Mincer-like specification**: `age`, `age^2`, `homme`,
    `education`, `region`. `age` proxies experience; the quadratic
    captures the hump-shaped earnings profile. `region` controls for
    geographic wage variation. We use `education` (six levels) rather
    than finer versions to keep enough observations in each level.
-   **(3)/(7) + background**: `expe_specifique` (firm tenure),
    `csp_pere_gp` and `csp_mere_gp` (parental CSP, 1-digit), `matri`,
    `sante_declaree`, `tuu_r`, `zus` (individual and family
    characteristics). `zus` captures spatial segregation (correlated
    with both origin and wages). `sante_declaree` is a productivity
    proxy (though potentially endogenous via reverse causality).
-   **(4)/(8) + job characteristics**: `csp_actif`, `contrat`,
    `typehoraire`. However, these are partly determined by origin, so we
    include them only to study the channels through which origin affects
    wages (Question 1c), not in our preferred specification.

Below is the detailed explanation for the variables we chose to include
or not in our models.

#### 1. Included Variables

**Demographic and geographic controls.** We control for `age` (using a
quadratic form to capture the concave, hump-shaped earnings profile
typical of a life-cycle career) and the gender indicator `homme` to
account for experience and the gender wage gap. Geographic variables
(`region`, `tuu_r`, and `zus`) are crucial to isolate origin from
spatial segregation, varying local labor markets, and metropolitan
cost-of-living differences. Legal marital status (`matri`) is added as
it often signals stability and correlates with higher earnings in the
academic literature. Lastly, self-reported health (`sante_declaree`)
acts as a proxy for individual productivity, despite the fact that there
might be some potential endogeneity from reverse causality.

**Family background and education.** The `education` variable is a
crucial control to separate the direct effect of origin from educational
attainment. To control for social reproduction and intergenerational
wealth or status transfers, we include parental occupations (`csp_pere`
and `csp_mere`). We agregate these categories (`csp_pere_gp`,
`csp_mere_gp`) to gain statistical power.

**Job characteristics.** We eventually include `expe_specifique` to
compare workers at equal levels of firm tenure. The `plract` indicator
controls for multiple job-holding, which may signal financial necessity.
Variables like `typehoraire` and `compenheuresup` account for
compensating differentials like atypical hours or overtime pay. However,
while variables like `csp_actif` and `contrat` are included in extended
specifications, they are partly determined by origin itself. Including
them absorbs the indirect effect of origin, which is why they are used
exclusively in Question 1c, not in our preferred baseline.

#### 2. Excluded Variables

**Labor status and mechanical collinearity.** Variables related to
unemployment or inactivity (`acteu`, `chomeur`, and `ancchom`) are
excluded because these individuals are not relevant in our sample.
Similarly, `heuretra`, `jourtra`, and `salmensuel` are omitted to avoid
mechanical collinearity, as the dependent variable (`salhoraire`) is
computed directly from them. Finally, `taille_menage` is excluded due to
a high non-response bias across wage data and the absence of a clear
theoretical link to hourly wages.

**Collinearity and redundancy.** Parental nationalities (`nat_mere`,
`nat_pere`) are near-collinear with the `origine` variable (which is
constructed directly from them), so they are dropped. Individual
nationality (`nat`) is also omitted to strictly isolate the effect of
origin rather than legal citizenship. We exclude `pacs` in favor of
`matri` because `pacs` contains thousands of missing values.
Furthermore, `descimmi` and `immi` are excluded since their information
is already largely captured by the origin categories themselves.

### Functional form

```{r}
# Age vs log wage with quadratic fit
p1 <- ggplot(df_q1, aes(age, logsalhoraire)) +
  geom_point(alpha = 0.15, size = 0.8, colour = "#248BFF") +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2),
              colour = "#e81757", linewidth = 0.8) +
  labs(title = "Age vs log hourly wage",
       x = "Age", y = "Log hourly wage") +
  theme_projet()

# Firm tenure vs log wage
p2 <- ggplot(df_q1, aes(expe_specifique, logsalhoraire)) +
  geom_point(alpha = 0.15, size = 0.8, colour = "#248BFF") +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2),
              colour = "#e81757", linewidth = 0.8) +
  labs(title = "Firm tenure vs log hourly wage",
       x = "Firm tenure (months)", y = "Log hourly wage") +
  theme_projet()

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

The scatter plots confirm a concave age--wage profile, justifying the
quadratic specification `age + age^2`. The relationship with firm tenure
is approximately linear, so we include `expe_specifique` without a
quadratic term.

### Why we pick the log specification

We prefer the log specification necause coefficients are directly
interpretable as approximate percentage effects, which is more natural
than a fixed euro amount. Moreover, since AIC is not directly comparable
across different dependent variables, we report the level specification
as a robustness check.

Among log specifications, our preferred model is **(7)**: it has the
highest adjusted R-squared among specifications without bad controls,
and all its covariates have clear theoretical justification.

## 1b) Interpretation of the origin coefficients

```{r}
# Comment: We compute the mean French-origin wage as a reference point,
# then extract the origin coefficients from specification (7) and convert
# them to percentage effects and euro amounts.
mean_fr <- mean(df_q1$salhoraire[df_q1$origine == "France"], na.rm = TRUE)
cat("Mean hourly wage for French-origin individuals:", round(mean_fr, 2), "EUR\n")

coef_tab <- broom::tidy(reg7_log, conf.int = TRUE) %>%
  filter(grepl("^origine", term)) %>%
  mutate(
    pct_effect = (exp(estimate) - 1) * 100,
    euro_at_mean = pct_effect / 100 * mean_fr
  )
coef_tab %>%
  select(term, estimate, std.error, p.value, conf.low, conf.high,
         pct_effect, euro_at_mean)
```

In the preferred log specification **(7)**, each coefficient on
`origine` is interpreted as a semi-elasticity: it gives the approximate
percentage wage gap between the corresponding origin group and
individuals of French origin, holding age, sex, education, region,
firm-specific experience, parental CSP, marital status, self-reported
health, urban unit, and ZUS status fixed. For small coefficients, the
approximation $\hat\beta \approx$ percentage effect holds. For larger
coefficients, the exact formula $(\exp(\hat\beta) - 1) \times 100$ is
more accurate.

In all our specifications, the origin coefficients strongly depend on
the modality. Compared to individuals of French origin:

-   Some European-origin groups show a positive wage premium (e.g.
    Southern Europe, Northern Europe), which remains significant at the
    5% level in several specifications.
-   Other groups (notably Rest of Africa) show a negative wage gap,
    though its significance varies across specifications.
-   Most other modalities (Eastern Europe, Maghreb, Southeast Asia, Rest
    of the world, Unknown) have point estimates between -5% and +2%, and
    are not individually significant at conventional levels.

However, non-significance for small subgroups (often fewer than 200
observations) should not be interpreted as evidence of no gap: it may
simply reflect low statistical power.

## 1c) Mechanisms

It is likely that the effect of origin on wages is partly mediated
through other variables. To test this, we start from the Mincer baseline
(specification **(6)**:
`origin + age + age² + sex + education + region`) and progressively add
each candidate channel, tracking how the origin coefficients move. The
key channels we investigate are: parental social background
(`csp_pere_gp`), occupation (`csp_actif`), type of contract (`contrat`),
and type of working hours (`typehoraire`).

```{r}
m_base <- lm(logsalhoraire ~ origine + age + I(age^2) + homme + region,
             data = df_q1)

chan_specs <- list(
  "baseline (no educ)" = m_base,
  "+ education"        = update(m_base, . ~ . + education),
  "+ csp_pere_gp"      = update(m_base, . ~ . + education + csp_pere_gp),
  "+ csp_actif"        = update(m_base, . ~ . + education + csp_actif),
  "+ contrat"          = update(m_base, . ~ . + education + contrat),
  "+ typehoraire"      = update(m_base, . ~ . + education + typehoraire),
  "all channels"       = reg8_log
)

# Comment: Collect origin coefficients from each specification and plot
coef_evo <- map_dfr(chan_specs, ~ broom::tidy(.x), .id = "spec") %>%
  filter(grepl("^origine", term)) %>%
  mutate(spec = factor(spec, levels = names(chan_specs)))

ggplot(coef_evo,
       aes(x = spec, y = estimate, colour = term, group = term)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_line() +
  geom_point() +
  coord_flip() +
  labs(x = NULL,
       y = "Coefficient on log(hourly wage)",
       title = "Evolution of origin coefficients as control variables are added") +
  theme_projet()
```

```{r}
modelsummary(
  chan_specs,
  vcov = vcov_robust,
  stars = TRUE, output = "markdown",
  coef_omit = "^\\(Intercept\\)|^education|^region|^csp_pere_gp|^csp_actif|^contrat|^typehoraire|^matri|^sante_declaree|^tuu_r|^zus|^csp_mere_gp|^expe_specifique",
  title = "Sequential mediation analysis (log hourly wage)",
  gof_omit = "IC|Log|F|RMSE"
)
```

The plot and table make it clear which channels absorb the origin gap:

-   **Education**: adding education as a control reduces the negative
    coefficients for some origin groups (notably Rest of Africa), which
    is consistent with lower average educational attainment mediating
    part of the wage gap for these groups.

-   **Occupation (`csp_actif`)**: this is the single most powerful
    channel. Adding it typically halves the origin gap, consistent with
    origin groups being over-represented in lower-paying occupations.
    This does not mean origin has no effect on wages but that the effect
    largely operates through occupational sorting.

-   **Contract type (`contrat`)**: controlling for contract type moves
    some coefficients further, pointing to a segmentation of labor
    contracts by origin (e.g. more fixed-term contracts for some
    groups).

-   **Parental CSP (`csp_pere_gp`)**: absorbs part of the gap for
    second-generation groups, consistent with intergenerational
    transmission of social status.

# Question 2: Education level

## 2a) Education as a problematic control

We can argue that education is partly explained by individual and family
choices that themselves depend on origin (e.g. cultural context, access
to information on education, resources, etc.). In other words, origin
impacts wages largely indirectly through its impact on education, as
shown above. By adding education to the model, the estimated coefficient
for origin only captures the direct effect on wages, ignoring the
indirect effect through educational pathways that may be created by
unobservable factors. Consequently, we can expect the coefficient for
origin to become smaller in absolute value than the total effect, as it
no longer accounts for the potential disadvantages accumulated during
the individual's educational trajectory.

(An other phenomenon, althought that's not really our interest in this
question, is that the coefficient for education may be biased due to
omitted unobservable factors (such as innate talent or motivation) that
simultaneously influence both educational level and wages).

## 2b) Instruments

To address the main problem mentioned in (2a), we can use an
instrumental variable.

For a variable $Z$ to be a valid instrument, it must satisfy two
conditions:

1.  Relevance: $Cov(\text{educ}, Z) \neq 0$. The instrument must be
    strongly correlated with education.

2.  Exogeneity: $Cov(\text{salhoraire}, Z) = 0$. $Z$ should not be
    correlated with the error term but only impact wages indirectly
    through education.

By intuition, `csp_mere` or `csp_pere` seem to be good choices:

1.  **Relevance**: Parental socio-professional category influences a lot
    the willingness and capacity to pursue long studies, often through
    social reproduction or different visions of job possibilities.

2.  **Exogeneity**: Parental CSP impacts wages primarily indirectly.
    Even if it shapes educational choices and cultural capital, an
    individual's current hourly wage is mostly determined by their own
    qualifications and career path, making any remaining direct
    influence from parents' status possibly negligible.

To the contrary, `nat_mere` and `nat_pere` are not valid instruments:
`origine` is constructed from parental nationalities so they are nearly
collinear with the exogenous regressor `origine` that we want to keep in
the second stage. Using them as instruments would violate the exclusion
restriction.

To identify the causal effect of one exogeneous variable, we need one
valid instrument (one instrument for each exogeneous variable of our
model). Since we treat education as a categorical variable, each level
(represented by a dummy variable) is potentially endogenous. Therefore,
we need at least as many instruments as there are education levels
(minus the reference category). Using `csp_mere` as a categorical
instrument, for instance, provides enough instruments to identify the
effect of each education level (which is also the case for
`nat_mere`/`nat_pere`)

However, the `csp_mere` and `csp_pere` variables have too many
categories. For the instruments to be robust, each category must contain
a significant number of individuals. To achieve this, we group them by
"Groupe Socioprofessionnel" (the 1st digit level) according to the INSEE
nomenclature. This grouping ensures we have enough instruments to
satisfy the order condition while maintaining statistical power.

```{r}
df_q1 <- df_q1 %>%
  mutate(
    # Extract the first digit of parental CSP
    csp_pere_gp = substr(as.character(csp_pere), 1, 1),
    csp_mere_gp = substr(as.character(csp_mere), 1, 1)
  ) %>%
  mutate(
    # Convert to factors for the regression
    csp_pere_gp = as.factor(csp_pere_gp),
    csp_mere_gp = as.factor(csp_mere_gp)
  )

summary(df_q1$csp_mere_gp)
```

We have 8 instruments \> 5 exogenous variables (6 levels of education so
5 dummies in the model).

## 2c) 2SLS estimation

We estimate the 2SLS using the same controls as our preferred OLS
specification **(7)**, so that the comparison between OLS and IV is
meaningful.

```{r}
# Instrument set 1: csp_pere_gp only
model_iv_1 <- ivreg(
  logsalhoraire ~ education + origine + age + I(age^2) + homme + region +
    expe_specifique + csp_mere_gp + matri + sante_declaree + tuu_r + zus |
    csp_pere_gp + origine + age + I(age^2) + homme + region +
    expe_specifique + csp_mere_gp + matri + sante_declaree + tuu_r + zus,
  data = df_q1)

# Instrument set 2: csp_pere_gp + csp_mere_gp
model_iv_2 <- ivreg(
  logsalhoraire ~ education + origine + age + I(age^2) + homme + region +
    expe_specifique + matri + sante_declaree + tuu_r + zus |
    csp_pere_gp + csp_mere_gp + origine + age + I(age^2) + homme + region +
    expe_specifique + matri + sante_declaree + tuu_r + zus,
  data = df_q1)

summary(model_iv_1, diagnostics = TRUE)
summary(model_iv_2, diagnostics = TRUE)
```

```{r}
modelsummary(
  list("OLS (7)" = reg7_log,
       "2SLS (csp_pere)" = model_iv_1,
       "2SLS (csp_pere + csp_mere)" = model_iv_2),
  vcov = list(vcov_robust, vcov_robust, vcov_robust),
  stars = TRUE, output = "markdown",
  coef_omit = "^\\(Intercept\\)|^region|^matri|^sante_declaree|^tuu_r|^zus|^csp_pere_gp|^csp_mere_gp",
  title = "OLS vs 2SLS --- origin and education",
  gof_omit = "IC|Log|F|RMSE"
)
```

## 2d) Interpretation

We can make three comments here:

1.  **Diagnostics.** The first-stage *F*-statistic on the excluded
    instruments should exceed the Stock--Yogo critical value (rule of
    thumb: *F* \> 10) for the instruments to be considered strong. If
    the Sargan over-identification test rejects, at least one instrument
    is correlated with the wage residual and we should prefer the
    just-identified specification using only `csp_pere_gp`.

2.  **Education coefficient.** We expect the IV coefficient on education
    dummies to be larger in absolute value than OLS if the ability bias
    was attenuating the OLS estimate, but also noisier due to the loss
    of efficiency from 2SLS.

3.  **Origin coefficient.** If the IV origin coefficients remain
    essentially unchanged compared to OLS after instrumenting education,
    it means the indirect-through-education channel is small and the
    bulk of the origin gap passes through other channels (occupation,
    contract type, parental networks). This is consistent with the
    mediation analysis of Question 1c. The wider standard errors are
    mechanical (2SLS is less efficient than OLS), so non-significance of
    individual origin modalities should not be over-interpreted; 95%
    confidence intervals are more informative than *p*-values here.

# Question 3: Panel data

## 3a) 
The wording tells us that we first assume to have the exogeneity of residuals and individual fixed effect. The exogeneity of individual fixed effect means that we can separate the effect of variables that are time invariant like origin from the individual residual alpha_i.
As we need to use the panel structure at trimester 1 and 6 we can use the model 7 of Q1 to which we add an indicator of being in period t=1 or t=6 (for R to identify the panel structure we need to use plm instead of lm).

```{r}
#to do so we first need to clean the needed datas and specify it is a panel data with periods t through pdata.frame
df_q3 <- df %>%   filter(t %in% c(1, 6), acteu == "1", !is.na(salhoraire)) %>%   filter(complete.cases(select(., salhoraire, logsalhoraire, origine, age, homme, education, region, tuu_r, zus, matri, sante_declaree, csp_pere_gp, csp_mere_gp,     expe_specifique, id_menage))) %>%   arrange(id_individu, t)  
pdata_q3 <- pdata.frame(df_q3, index = c("id_individu", "t"))  

# And here is the regression with indicators for periods
re_q3 <- plm( logsalhoraire ~ origine + age + I(age^2) + homme + education + region + expe_specifique + csp_pere_gp + csp_mere_gp + matri + sante_declaree + tuu_r + zus + factor(t),   data  = pdata_q3,   model = "random",   effect = "individual" ) 
coeftest(re_q3, vcovHC(re_q3, cluster = "group", type = "HC1"))
```
The factor(t)6 coefficient (+0.014, significant at 1%) indicates average hourly wages grew by roughly +1.4% between the first and last observed quarters (2014–2016).

Individual wage profile: The age quadratic confirms the expected concave life-cycle for wages, peaking at around 52 years old (Beta_age_hat + 2 x Beta_square_age_hat x age = 0 implies age=52~).
Men earn about +17.4% more than women (exp(0.160) − 1), all else equal.
Education: The gradient is strong and monotonic: compared to the reference category (higher education beyond bac+2), wages fall by 15% at bac+2 level, 23% at baccalauréat, and up to −39% ((exp(−0.496)−1)×100 formula) for individuals with no qualification — all highly significant.

Origin: Only two groups show statistically significant wage gaps relative to French-origin workers, controlling for all observables:

Southern Europe: +4.9% wage premium (significant at 5%), consistent with positive selection of this group into well-paying sectors.
Rest of Africa: −10.0% wage penalty (significant at 5%), the only significant negative gap, suggesting persistent wage inequality for this group even after controlling for education, region, occupation and family background.

All other origin groups yield insignificant coefficients, likely reflecting low statistical power due to small subgroup sizes rather than the absence of any gap. A joint Wald test on all origin dummies remains necessary to assess the overall significance of the variable.
```{r}

names_origin <- grep("origine", names(coef(re_q3)), value = TRUE)


wald_test <- linearHypothesis(re_q3, names_origin, 
                              vcov = vcovHC(re_q3, method = "arellano", cluster = "group"))

print(wald_test)
```
The Test rejects the joint nullity of the coefficient at the 5% level, thus in this model the origin have significant impact on individual's wages.
Comparison with Q1. Results are qualitatively consistent with the cross-sectional specification (7). The "random effect" panel adds gains from exploiting both within and between-individual variation and partially accounts for time-invariant unobserved heterogeneity  — subject to the exogeneity assumption $E[\alpha_i|X]=0$ discussed in 3b).

##3b)




# Question 4: Linear probability model on unemployment

## Sample and dependent variable

In this question, we focus on active individuals (`acteu == "1"` or
`acteu == "2"`) across all six quarters and denote by $\alpha_i$ the
unobserved, time-invariant individual heterogeneity. `sante_declaree` is
only observed at $t = 1$ and $t = 6$, so we drop it from the regressors
list to avoid losing four quarters of observations.

```{r}
df_q4 <- df %>%
  filter(acteu %in% c("1", "2")) %>%
  # Dummy for unemployed individuals
  mutate(u = as.integer(acteu == "2")) %>%
  # Covariates
  select(id_individu, t, id_menage, u,
         origine, age, homme, education, region,
         matri, tuu_r, zus,
         csp_pere_gp, csp_mere_gp) %>%
  # Keep complete observations
  filter(complete.cases(.)) %>%
  arrange(id_individu, t)
pdata_q4 <- pdata.frame(df_q4, index = c("id_individu", "t"))
```

## 4a) Panel LPM

Even though the dependent variable is binary, we start by modeling it
using a Linear Probability Model:

$u_{it}=\mathbf{X}_{it}'\beta+\mathbf{Z}_i'\delta+\alpha_i+\varepsilon_{it}$

where $u_{it}$ is the binary dependent variable indicating unemployment
(1 if unemployed, 0 otherwise), $\mathbf{X}_{it}$ is a vector of
time-varying observable characteristics (age, household structure,
region etc.) and $\mathbf{Z}_i$ is a vector of time-invariant observable
characteristics (origin). We assume the unobserved individual
heterogeneity $\alpha_i$ is completely exogenous, meaning it is
uncorrelated with our explanatory variables.

Only age varies within individual. All other variables are
time-invariant over six quarters in this sample. Still, we include
quarter dummies in our Fixed effects model to correct for aggregate
shocks.

```{r}
fe_q4 <- plm(u ~ age + I(age^2) + factor(t), pdata_q4, model = "within", effect = "individual")
coeftest(fe_q4, vcovHC(fe_q4, cluster = "group", type = "HC1"))
```

If we include all the time-invariant controls (including `origine`), we
get our random effects model:

```{r}
re_q4 <- plm(u ~ origine + age + I(age^2) + homme + education + region + tuu_r + zus + matri + csp_pere_gp + csp_mere_gp + factor(t), pdata_q4, model = "random", effect = "individual")
coeftest(re_q4, vcovHC(re_q4, cluster = "group", type = "HC1")) %>%
  tidy() %>%
  filter(grepl("^origine|education|age", term))
```

Here, origin estimates fall between -0.04 (Rest of the world) and 0.10
(Maghreb). This means that having a Maghrebian origin is associated with
a 10 percentage points higher probability of being unemployed,
everything else observed being held equal.

To choose between our Fixed Effects (FE) and our Random Effects (RE)
models, we run a Hausman test on the comparable sub-model (time-varying
regressors only).

```{r}
re_q4_comp <- plm(u ~ age + I(age^2) + factor(t), pdata_q4, model = "random", effect = "individual")
phtest(fe_q4, re_q4_comp)
```

Here, our *p*-value is very close to 1, meaning that we fail to reject
the null hypothesis that the coefficients generated by the FE model and
the RE model for `age`, `age^2`, and the time dummies are practically
identical. This means that the data provides no evidence that $\alpha_i$
is correlated with age or time. Statistically, the exogeneity assumption
of the Random Effects model holds up for this specific specification.

However, fitted probabilities in a linear probability model may fall
outside [0, 1]. This is the reason why we also report a cross-sectional
probit/logit in Question 6.

## 4b) `origine` and the fixed-effects framework

If we believe unobserved characteristics (like motivation or family
network) are correlated with our regressors, we must use a Fixed Effects
(within) model which works by subtracting the individual's average over
time to eliminate $\alpha_i$.

Therefore, the same issue than in question 3c) arises if we allow for
correlation between individual effects and the regressors: `origine` is
time-invariant and therefore not identified by a within-FE estimator:
because an individual's origin is constant across all 6 quarters,
subtracting the individual mean completely drops this variable.

# Question 5: Dynamic model of unemployment state dependence

## 5a) Specification and exogeneity assumptions

In this question, we estimate the specification:
$u_{it} = \gamma u_{i,t-1} + \mathbf{X}_{it}' \beta + \alpha_i + \varepsilon_{it}$
for $t = 2,\ldots, 6$, on the sample of individuals active in every
quarter. Here, time-invariant controls $\mathbf{Z}_i$ are omitted since
Arellano-Bond estimators rely on first-differencing. In this context:

-   **Strict exogeneity**
    ($E[\varepsilon_it | u_{i,1...6}, \mathbf{X}_i, \alpha_i] = 0$ is
    not credible: a shock that increases `u_{it}` mechanically flows
    into `u_{it+1}` (which is a regressor in the next period), so the
    lagged dependent variable violates strict exogeneity by
    construction.

-   Instead, **weak exogeneity**
    ($E[\varepsilon_it | u_{i,1..t-1}, \mathbf{X}_{i,1..t}, \alpha_i] = 0$)
    is a more plausible assumption that justifies using the general
    method of moments (GMM), specifically the Arellano-Bond estimator.

This method takes the first difference of the equation to eliminate
$\alpha_i$ and then uses distant lags (e.g., $t−2$, $t−3$) of
unemployment as instruments for the differenced lagged dependent
variable.

```{r}
df_q5 <- df %>%
  filter(acteu %in% c("1", "2")) %>%
  mutate(u = as.integer(acteu == "2")) %>%
  group_by(id_individu) %>%
  filter(n() == 6) %>%
  ungroup() %>%
  select(id_individu, t, id_menage, u, age) %>%
  arrange(id_individu, t)
pdata_q5 <- pdata.frame(df_q5, index = c("id_individu", "t"))
```

```{r}
# Pooled OLS
ols_q5 <- plm(u ~ lag(u, 1) + age + I(age^2), pdata_q5, model = "pooling")

# Within estimator
fe_q5  <- plm(u ~ lag(u, 1) + age + I(age^2), pdata_q5, model = "within", effect = "individual")

# Arellano–Bond two-step GMM
ab_q5  <- pgmm(u ~ lag(u, 1) + age + I(age^2) | lag(u, 2:5), pdata_q5, effect = "individual", model = "twosteps", transformation = "d")
```

```{r}
bind_rows(
  summarise_gamma(ols_q5, "Pooled OLS"),
  summarise_gamma(fe_q5,  "Within"),
  summarise_gamma(ab_q5,  "Arellano–Bond GMM")
)
```

The pooled OLS yields $\hat \gamma_\text{Pooled OLS}=0.686$. Being
unemployed last quarter is associated with a 68.6 percentage point
increase in the probability of being unemployed this quarter. However,
this estimate is biased upward, as the lagged variable is mechanically
positively correlated with the unobserved individual heterogeneity.

The within estimator gives $\hat \gamma_\text{WE}=0.086$. Being
unemployed last quarter only increases the probability of being
unemployed this quarter by 8.6 percentage points. This estimate is
biased downward due to the Nickell bias as the within-transformation
introduces a negative correlation between the transformed lagged
dependent variable and the transformed error term.

Eventually, the Arellano-Bond GMM yields $\hat \gamma_\text{GMM}=0.380$,
which falls between the pooled OLS and the within estimator estimates,
purging both the upward heterogeneity OLS bias and the downward WE
Nickel bias. This means that experiencing unemployment last quarter is
associated with a 38 percentage points higher likelihood of being
unemployed this quarter everything else observed and unobserved held
equal.

We eventually run the Sargan/Hansen J-test for instrument validity and
the AR(2) test for serial correlation

```{r}
ab_sum <- summary(ab_q5)
ab_sum$m.tests   ## AR(1), AR(2)
ab_sum$sargan    ## Sargan / Hansen J
```

Since we get a *p*-value of 0.0323, we reject the null hypothesis that
instruments as a group are strictly exogenous at the standard 5%
significance level. This implies that our instruments might not be
entirely exogenous.

## 5b) Interpretation

The true state dependence parameter $\gamma$ is logically bounded by our
three estimates, which behave exactly as econometric theory predicts:

1)  The pooled OLS provides an upper bound. It is biased upward because
    it confounds true state dependence with unobserved individual
    heterogeneity ($\alpha_i$). It captures spurious state dependence,
    meaning part of the raw persistence we see is simply due to
    unobserved, time-invariant personal traits rather than the actual
    experience of being unemployed.
2)  The Within estimator provides a lower bound. It suffers from a
    downward Nickell bias, which is severe here given our short panel
    length of $T=6$ (yielding only 5 time periods for the lagged
    specification).
3)  The Arellano-Bond estimator is our preferred estimate. Falling
    between the pooled OLS and Within estimator estimates, it purges the
    individual fixed effects and bypasses the Nickell bias.

Consequently, the GMM estimate provides our preferred measure of state
dependence (the "scarring effect"). It shows that having been unemployed
last quarter raises the probability of being unemployed this quarter by
exactly $\hat{\gamma}_\text{GMM}=38$ percentage points, holding all
unobserved individual heterogeneity constant.

# Question 6: Probit / logit for unemployment

We finally address the binary nature of the dependent variable properly
by moving to non-linear models.

## 6a) Within and first difference for probit/logit?

Non-linear panel models do not admit a transformation that cleanly
considers the individual effect $\alpha_i$:

-   The within transformation of a probit does not yield a standard
    probit as the link function is not additively separable.
-   Including individual dummies triggers the incidental-parameters
    problem: with $T$ small, the MLE of $\beta$ is inconsistent because
    the number of nuisance parameters grows with $N$.

## 6b) Cross-sectional probit and logit at $t=6$

We collapse the data to a cross-section using only the final quarter and
estimate standard probit and logit models.

```{r}
df_q6 <- df %>%
  filter(acteu %in% c("1", "2"), t == 6) %>%
  mutate(u   = as.integer(acteu == "2"),
         hed = as.integer(education == "1")) %>%
  select(id_individu, id_menage, u, hed,
         origine, age, homme, education, region,
         tuu_r, zus, matri, sante_declaree,
         csp_pere_gp, csp_mere_gp) %>%
  filter(complete.cases(.))
```

```{r}
f_q6 <- u ~ origine + age + I(age^2) + homme + education + region +
  tuu_r + zus + matri

probit_q6 <- glm(f_q6, data = df_q6, family = binomial("probit"))
logit_q6  <- glm(f_q6, data = df_q6, family = binomial("logit"))
```

Probit uses a standard normal distribution, while logit uses a logistic
distribution. Because their scales differ, their raw coefficients are
meaningless to compare directly. Instead, we compute and compare their
Average Marginal Effects (AME) using `margins::margins()`. The AMEs
provide the actual average percentage-point impact of "origin" on the
probability of being unemployed.

```{r}
ame_probit <- summary(margins(probit_q6, variables = "origine"))
ame_logit  <- summary(margins(logit_q6,  variables = "origine"))

ame_compare <- tibble(
  factor     = ame_probit$factor,
  AME_probit = ame_probit$AME,
  SE_probit  = ame_probit$SE,
  AME_logit  = ame_logit$AME,
  SE_logit   = ame_logit$SE
)
ame_compare
```

As expected, the Probit and Logit AMEs are numerically almost identical,
confirming that the choice of link function matters very little for
marginal effects.

Each row in the table gives the average effect of belonging to a
specific origin relative to the reference category (French origin),
holding age, sex, education, region, and other observables constant.

-   **Significant penalties**. We observe substantial and statistically
    significant employment penalties for certain origins. For example,
    looking at the Probit AME, individuals of Maghreb origin face a 9.8
    pp higher probability of being unemployed compared to individuals of
    French origin. Similarly, individuals of Eastern European origin
    face an 8.8 percentage point higher likelihood of unemployment.

-   **Insignificant differences.** Conversely, the AMEs for origins such
    as Northern Europe (-0.01 pp) or Southern Europe (+0.8 percentage
    points) are close to zero and non significant. This indicates no
    statistically significant difference in the probability of
    unemployment for these groups relative to individuals of French
    origin.

## 6c) Rivers--Vuong for endogenous higher education

We now let `hed = 1{education == 1}` be endogenous and use again the
agregated parental CSP as instrument. We also consider other exogenous
regressors: `origine`, `age`, `age^2`, `homme`, `reg`, urban unit, ZUS,
`matri`.

```{r}
fs_q6  <- lm(hed ~ csp_pere_gp + csp_mere_gp + origine + age + I(age^2) + homme + region + tuu_r + zus + matri, data = df_q6)

## joint F-statistic on the instruments (first-stage strength)
linearHypothesis(fs_q6,
                 grep("^csp_(pere|mere)_gp", names(coef(fs_q6)), value = TRUE),
                 vcov. = vcovHC(fs_q6, type = "HC1"))
```

The joint *F*-statistic on the excluded instruments in the first-stage
linear regression is 50.356 (*p*-val. \< 0.001). Because this is well
above the standard rule-of-thumb threshold of 10, we can confidently
conclude that our instruments are "strong" and do not suffer from the
weak instrument problem.

In a second stage, we run a probit of unemployment on `hed`, the
exogenous controls, and the first-stage residual $\hat v_i$.

```{r}
df_q6$v_hat <- residuals(fs_q6)

f_rv <- u ~ hed + origine + age + I(age^2) + homme + region +
  tuu_r + zus + matri + v_hat
rv_q6 <- glm(f_rv, data = df_q6, family = binomial("probit"))
coef(summary(rv_q6))[c("hed", "v_hat"), ]
```

The coefficient on $\hat v_i$ is positive but statistically
insignificant (*p*-val.=0.143) such that we fail to reject the null
hypothesis that higher education is exogenous. This means that once we
control for our set of observables, unobserved factors like ability or
motivation are not significantly biasing our estimate of the education
effect. In this specific sample, treating higher education as exogenous
is statistically justifiable.

Because we used a two-step procedure, we rely on the bootstrapped
standard errors for inference.

```{r}
rv_stat <- function(data, idx) {
  d <- data[idx, , drop = FALSE]
  fs <- lm(hed ~ csp_pere_gp + csp_mere_gp + origine + age + I(age^2) + homme + region + tuu_r + zus + matri, data = d)
  d$v_hat <- residuals(fs)
  m <- glm(f_rv, data = d, family = binomial("probit"))
  coef(m)
}
set.seed(42)
bs_rv <- boot(data = df_q6, statistic = rv_stat, R = 300)

boot_se <- apply(bs_rv$t, 2, sd, na.rm = TRUE)
names(boot_se) <- names(bs_rv$t0)

rv_table <- tibble(
  term        = names(bs_rv$t0),
  estimate    = bs_rv$t0,
  boot_se     = boot_se,
  z           = bs_rv$t0 / boot_se,
  p_value     = 2 * pnorm(-abs(bs_rv$t0 / boot_se))
) %>% filter(grepl("^origine|hed|v_hat", term))
rv_table
```

The coefficient on higher education remains negative (-0.621) and highly
significant (*p*-val.=0.008). As expected, obtaining a higher education
degree strongly protects against unemployment.

Eventually, we compare Average Marginal Effects (AME) under the naïve
probit and Rivers-Vuong.

```{r}
ame_rv <- summary(margins(rv_q6, variables = "origine"))

tibble(
  factor       = ame_probit$factor,
  AME_probit   = ame_probit$AME,
  AME_rv       = ame_rv$AME
)
```

The penalties associated with origin remain broadly stable. For
instance, the penalty for Maghreb origin shifts only slightly from 9.8
to 11.3 percentage points, and Eastern Europe remains at 8.8 percentage
points. This stability confirms that the employment penalties faced by
these groups are not a spurious artifact of endogenous sorting into
higher education. The ethnic penalties persist independently of the
education channel.
