---
title: "Project Econometrics"
author: "Emile Guillaume, Jean Lavallée and Aurélian Moge"
date: "2026-04-15"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      cache = TRUE, fig.width = 7, fig.height = 4)
```

# Packages, figure parameters, functions and data

```{r, include=FALSE}
pkgs <- c("haven", "tidyverse", "sjlabelled", "modelsummary",
          "AER", "sandwich", "lmtest", "car", "plm", "Formula", "broom",
          "gridExtra")
for (p in pkgs) if (!requireNamespace(p, quietly = TRUE)) install.packages(p)
library(haven); library(tidyverse); library(sjlabelled)
library(modelsummary); library(AER); library(sandwich)
library(lmtest); library(car); library(Formula); library(broom)
library(gridExtra)

# Function for figures theme
theme_projet <- function() {
  theme_bw() +
    theme(
      legend.position    = "bottom",
      legend.spacing.y   = unit(0.1, "cm"),
      legend.key.height  = unit(0.2, "cm"),
      legend.key.width   = unit(1,   "cm"),
      panel.border       = element_blank(),
      axis.line          = element_line(color = "black"),
      panel.grid.major.x = element_blank(),
      panel.grid.minor   = element_blank()
    )
}

# Function for household-clustered robust (HC1) variance
vcov_robust <- function(m) {
  vcovCL(m, cluster = ~ id_menage, type = "HC1")
}

# Function for robust Wald test that all origin dummies are jointly zero
wald_test_origin <- function(m) {
  nms <- grep("^origine", names(coef(m)), value = TRUE)
  if (length(nms) == 0) return(NA_real_)
  test <- linearHypothesis(m, nms, vcov. = vcov_robust(m),
                           # Under a robust vcov we use the Chi-square
                           test = "Chisq")
  test[2, "Pr(>Chisq)"]
}
```

```{r}
df <- read_dta("DM_Subject_2_Data.dta") %>%
  mutate(
    # Factor variables
    origine        = factor(origine),
    immi           = factor(immi),
    descimmi       = factor(descimmi),
    homme          = factor(homme),
    education      = factor(education),
    csp_actif      = factor(csp_actif),
    matri          = factor(matri),
    zus            = factor(zus),
    tuu_r          = factor(tuu_r),
    region         = factor(region),
    sante_declaree = factor(sante_declaree),
    contrat        = factor(contrat),
    typehoraire    = factor(typehoraire),
    compenheuresup = factor(compenheuresup),
    csp_pere_gp    = factor(substr(as.character(csp_pere), 1, 1)),
    csp_mere_gp    = factor(substr(as.character(csp_mere), 1, 1)),
    # Correct the "salhoraire" variable
    salhoraire     = salmensuel/(heuretra * 4.33),
    # Replace numeric codes with readable labels
    origine        = fct_recode(origine,
                       "France"                  = "01",
                       "Northern Europe"         = "03",
                       "Southern Europe"         = "04",
                       "Eastern Europe"          = "05",
                       "Maghreb"                 = "06",
                       "Rest of Africa"          = "07",
                       "Middle East"             = "08",
                       "Southeast Asia"          = "09",
                       "RoW"                     = "10",
                       "Unknown"                 = "99"
                     ),
    # France as reference
    origine        = relevel(origine, ref = "France")
  )
```

# Question 1: Cross-section of wages for employed individuals at t = 6

## Sample

In the general case, the wage is registered for the first and the last
quarters (T=1,2), among six quarters (t=1,2,3,4,5,6). Here, we restrict
our sample to employed individuals (i.e., `acteu == "1"`) in the last
quarter (i.e., T=2 or `t==6`), for whom hourly wage is available (i.e.,
`!is.na(salhoraire)`).

```{r}
df_q1 <- df %>% filter(t==6, acteu == "1", !is.na(salhoraire))
```

We also inspect individuals interviewed only twice (`rim` indicator, cf.
hint in the appendix).

```{r sample}
with(df_q1, table(rim, cut(age, c(-Inf, 30, 50, 65, Inf))))
```

Eventually, we restrict our sample to observations where no covariate is
missing

```{r}
covariates <- c("salhoraire", "logsalhoraire",
                 "origine", "immi", "age", "homme", "education",
                 "region", "tuu_r", "zus", "matri", "sante_declaree",
                 "csp_actif", "contrat", "typehoraire",
                 "csp_pere_gp", "csp_mere_gp",
                 "expe_specifique", "id_menage")
mask <- complete.cases(df_q1[, covariates])
df_q1 <- df_q1[mask, ]
nrow(df_q1)
```

After filtering observations, we run some descriptive statistics to get
a good understanding of the dataset.

```{r desc_stats, fig.width=10, fig.height=8}
# Origin
df_q1 %>%
  count(origine) %>%
  mutate(origine = fct_reorder(origine, n)) %>%
  ggplot(aes(x = origine, y = n)) +
  geom_col(fill = "#248BFF") +
  geom_text(aes(label = n), hjust = -0.15, size = 3.2) +
  coord_flip(clip = "off") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Distribution of employed individuals by origin (t=6)",
       x = "Origin", y = "Count") +
  theme_projet()

# Wages
df_q1 %>%
  ggplot(aes(x = logsalhoraire)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50,
                 fill = "#248BFF", colour = "white") +
  geom_density(colour = "#e81757", linewidth = 0.8) +
  labs(title = "Distribution of log hourly wages (t=6)",
       x = "Log hourly wage", y = "Density") +
  theme_projet()

# Log wages by origin
df_q1 %>%
  ggplot(aes(x = origine, y = logsalhoraire)) +
  geom_boxplot(fill = "#248BFF", colour = "#248BFF", alpha=0.2, outlier.size = 0.8) +
  labs(title = "Log hourly wages by origin (t=6)",
       x = "Origin", y = "Log hourly wage") +
  theme_projet()
```

The descriptive statistics reveal two key features of the data. First,
the origin distribution is skewed: the vast majority of employed
individuals at t=6 are of French origin, with relatively small subgroups
for each foreign origin. This imbalance will affect the precision of the
origin coefficients. Second, the distribution of log hourly wages is
approximately normal, supporting the use of a log-level specification.
The boxplots show moderate variation in median wages across origin
groups, but with substantial overlap, suggesting that any origin gap
must be carefully disentangled from other determinants.

## 1a) Estimation of the effect of some variables on wages

### First results

To begin with, we try a few specifications were we regress `salhoraire`
and `logsalhoraire` on variables that are commonly used in the
literature to explain wages, i.e., sequentially: age, gender, education,
location and socio-professional category. We also include `origin` in
all models as it is the independent variable we are interested in.

We first run specifications in levels

```{r}
# (1) Origin only
reg1_lvl <- lm(salhoraire ~ origine, df_q1)

# (2) Mincer-like specification = (1) + Mincer covariates
reg2_lvl <- lm(salhoraire ~ origine + age + I(age^2) + homme + education + region, df_q1)

# (3) = (2) + individual and family backgrounds
reg3_lvl <- lm(salhoraire ~ origine + age + I(age^2) + homme + education + region + expe_specifique + csp_pere_gp + csp_mere_gp + matri + sante_declaree + tuu_r + zus, df_q1)

# (4) = (3) + job characteristics
reg4_lvl <- lm(salhoraire ~ origine + age + I(age^2) + homme + education + region + expe_specifique + csp_pere_gp + csp_mere_gp + matri + sante_declaree + tuu_r + zus + csp_actif + contrat + typehoraire, df_q1)
```

And then use specifications in log

```{r}
# (5) Origin only
reg5_log <- lm(logsalhoraire ~ origine, df_q1)

# (6) Mincer-like specification = (5) + Mincer covariates
reg6_log <- lm(logsalhoraire ~ origine + age + I(age^2) + homme + education + region, df_q1)

# (7) = (6) + individual and family backgrounds
reg7_log <- lm(logsalhoraire ~ origine + age + I(age^2) + homme + education + region + expe_specifique + csp_pere_gp + csp_mere_gp + matri + sante_declaree + tuu_r + zus, df_q1)

# (8) = (7) + job characteristics
reg8_log <- lm(logsalhoraire ~ origine + age + I(age^2) + homme + education + region + expe_specifique + csp_pere_gp + csp_mere_gp + matri + sante_declaree + tuu_r + zus + csp_actif + contrat + typehoraire, df_q1)
```

We display the results in a single table

```{r display, results='asis'}
models <- list(
  "(1)" = reg1_lvl,
  "(2)" = reg2_lvl,
  "(3)" = reg3_lvl,
  "(4)" = reg4_lvl,
  "(5)" = reg5_log,
  "(6)" = reg6_log,
  "(7)" = reg7_log,
  "(8)" = reg8_log
  )

rows <- tribble(
  ~term,              ~"(1)", ~"(2)", ~"(3)", ~"(4)", ~"(5)", ~"(6)", ~"(7)", ~"(8)",
  "Education FE",     "",   "Yes",  "Yes",  "Yes",  "",   "Yes",  "Yes",  "Yes",
  "Region FE",        "",   "Yes",  "Yes",  "Yes",  "",   "Yes",  "Yes",  "Yes",
  "Matri FE",         "",   "",   "Yes",  "Yes",  "",   "",   "Yes",  "Yes",
  "Santé déclarée FE","",   "",   "Yes",  "Yes",  "",   "",   "Yes",  "Yes",
  "TUU_r FE",         "",   "",   "Yes",  "Yes",  "",   "",   "Yes",  "Yes",
  "ZUS FE",           "",   "",   "Yes",  "Yes",  "",   "",   "Yes",  "Yes",
  "CSP père FE",      "",   "",   "Yes",  "Yes",  "",   "",   "Yes",  "Yes",
  "CSP mère FE",      "",   "",   "Yes",  "Yes",  "",   "",   "Yes",  "Yes",
  "CSP actif FE",     "",   "",   "",   "Yes",  "",   "",   "",   "Yes",
  "Contrat FE",       "",   "",   "",   "Yes",  "",   "",   "",   "Yes",
  "Type horaire FE",  "",   "",   "",   "Yes",  "",   "",   "",   "Yes"
)

modelsummary(
  models,
  vcov = vcov_robust,
  stars = TRUE,
  output = "markdown",
  coef_omit = "^\\(Intercept\\)|^education|^region|^matri|^sante_declaree|^tuu_r|^zus|^csp_pere_gp|^csp_mere_gp|^csp_actif|^contrat|^typehoraire",
  title = NULL,
  gof_omit = "IC|Log|F|RMSE",
  add_rows = rows
  )
```

In all specifications, we report robust standard errors clustered at
the household level (`id_menage`).

We obtain models where some of the coefficients are statistically
significant at the 10%, 5% and 1% thresholds, and where the adjusted
R-squared falls between 0.004 (specification (1)) and 0.410
(specification (8)). We detail below our justification for each included
variable.

We also run a joint Wald test on origin dummies. Individual
origin coefficients are often non-significant due to small subgroup
sizes, but the joint test tells us whether origin as a whole is a
statistically meaningful determinant of wages. The joint Wald test
checks whether all origin dummies are simultaneously zero. This
is more informative than looking at individual p-values, especially
when subgroups are small as in this setting.

```{r wald_test, results='asis'}
wald_pvals <- tibble(
  Specification = names(models),
  `Wald p-value (origin)` = sapply(models, wald_test_origin)
)
knitr::kable(wald_pvals, digits = 3)
```

The Wald test of joint nullity of all origin dummies rejects in
specifications (1) and (5) (p-val < 0.01). As controls are added, the
p-value increases but typically remains below or near 0.05 (except for
specifications (3) and (4)), indicating that origin as a whole remains
a statistically meaningful determinant of wages even after controlling
for the usual covariates.

### Justification of control variables

Our goal is to obtain an unbiased estimator of the effect of `origine`
on wages. This requires $E[\varepsilon|\text{origine}] = 0$, meaning
there must be no omitted variables correlated with both origin and wages.
We therefore include controls that are plausibly correlated with both.

One thing we want when estimating the effect of origin on individual's                                        
wages by linear regression is to find an unbiased estimator of the                                            
parameter associated to origin in such a regression. For that to be                                           
verified, we need:
1.  The linearity in the parameter (we will specify our models to do so);
2.  the sample to be representative of the population (we have no hold
over it but we may discuss it later);
3.  the non-perfect colinearity of the independent variables we are going
to choose in our model (we will have to check it for each added variable);
4.  the nullity of the conditional expectation of the error term of the
regression ($E[\varepsilon|origin]=0$),
      
Our two minimal theoretical models are
$\log(salhoraire) = \beta_0 + \beta_1 origin + \varepsilon$ (column (1))
and
$salhoraire = \delta_0 + \delta_1 origin + \varepsilon$ (column (5))
      
Our four nested specifications progressively add:

-   **(1)/(5) Base model**: `origine` only --- the raw origin gap.
-   **(2)/(6) Mincer baseline**: `age`, `age^2`, `homme`, `education`,
    `region` --- standard wage determinants. `age` proxies experience;
    the quadratic captures the hump-shaped earnings profile.
    `region` controls for geographic wage variation. We use `education`
    (6 levels) rather than finer versions to keep cell sizes adequate.
-   **(3)/(7) + background**: `expe_specifique` (firm tenure),
    `csp_pere_gp` and `csp_mere_gp` (parental CSP, 1-digit),
    `matri`, `sante_declaree`, `tuu_r`, `zus` --- individual and family
    characteristics. `zus` captures spatial segregation (correlated with
    both origin and wages). `sante_declaree` is a productivity proxy
    (though potentially endogenous via reverse causality).
-   **(4)/(8) + job characteristics**: `csp_actif`, `contrat`,
    `typehoraire` --- these are "bad controls" (outcomes partly
    determined by origin), so we include them only to study the channels
    through which origin affects wages (Question 1c), not in our
    preferred specification.
    
Below is the detailed rationale for the variables we chose to include or exclude from our models.

#### 1. Included Variables

**Demographic & Geographic Controls:**
We control for `age` (using a quadratic form to capture the concave,
hump-shaped earnings profile typical of a life-cycle career) and the
gender indicator `homme` to account for experience and the gender wage
gap. Geographic variables (`region`, `tuu_r`, and `zus`) are crucial to
isolate origin from spatial segregation, varying local labor markets, and
metropolitan cost-of-living differences. Legal marital status (`matri`)
is added as it often signals stability and correlates with higher earnings.
Lastly, self-reported health (`sante_declaree`) acts as a proxy for
individual productivity, despite acknowledging potential endogeneity from
reverse causality.

**Family Background & Education:**
The `education` variable is a crucial control to separate the direct
effect of origin from educational attainment pathways. To control for
social reproduction and intergenerational wealth or status transfers, we
include parental occupations (`csp_pere` and `csp_mere`). We group these
into 1-digit categories (`csp_pere_gp`, `csp_mere_gp`) to maintain robust
statistical power and avoid volatile estimates from small cell sizes.

**Job Characteristics (Mediation Channels):**
We include `expe_specifique` to compare workers at equal levels of firm
tenure. The `plract` indicator controls for multiple job-holding, which
may signal financial necessity. Variables like `typehoraire` and
`compenheuresup` account for compensating differentials like atypical
hours or overtime pay. While variables like `csp_actif` and `contrat`
are included in extended specifications, we acknowledge them as "bad
controls" because occupational sorting and contract types are outcomes
partly determined by origin itself. Including them absorbs the indirect
effect of origin, which is why they are used exclusively to study mediation
channels (Question 1c), not in our preferred baseline.

#### 2. Excluded Variables

**Labor Status & Mechanical Collinearity:**
Variables related to unemployment or inactivity (`acteu`, `chomeur`,
and `ancchom`) are excluded because these individuals naturally have
no reported wages in our sample, rendering these variables statistically
irrelevant for the sub-population of employed workers. Similarly,
`heuretra`, `jourtra`, and `salmensuel` are omitted to avoid mechanical
collinearity, as the dependent variable (`salhoraire`) is computed directly
from them. Finally, `taille_menage` is excluded due to a high non-response
bias across wage data and the absence of a clear theoretical link to hourly
wages.

**Collinearity & Redundancy:**
Parental nationalities (`nat_mere`, `nat_pere`) are near-collinear
with the `origine` variable (which is constructed directly from them),
so they are dropped. Individual nationality (`nat`) is also omitted to
strictly isolate the effect of origin rather than legal citizenship. We
exclude `pacs` in favor of `matri` because `pacs` contains thousands of
missing values. Furthermore, `descimmi` and `immi` are excluded since their
information is already largely captured by the origin categories themselves.

### Functional form

```{r func_form, fig.width=10, fig.height=4}
# Age vs log wage with quadratic fit
p1 <- ggplot(df_q1, aes(age, logsalhoraire)) +
  geom_point(alpha = 0.15, size = 0.8, colour = "#248BFF") +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2),
              colour = "#e81757", linewidth = 0.8) +
  labs(title = "Age vs log hourly wage",
       x = "Age", y = "Log hourly wage") +
  theme_projet()

# Firm tenure vs log wage
p2 <- ggplot(df_q1, aes(expe_specifique, logsalhoraire)) +
  geom_point(alpha = 0.15, size = 0.8, colour = "#248BFF") +
  geom_smooth(method = "loess",
              colour = "#e81757", linewidth = 0.8) +
  labs(title = "Firm tenure vs log hourly wage",
       x = "Firm tenure (months)", y = "Log hourly wage") +
  theme_projet()

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

The scatter plots confirm a concave age--wage profile, justifying the
quadratic specification `age + age^2`. The relationship with firm tenure
is approximately linear, so we include `expe_specifique` without a
quadratic term.

### Why we pick the log specification

We prefer the log specification for two reasons: (i) coefficients are
directly interpretable as approximate percentage effects, which is more
natural than a fixed euro amount; (ii) AIC is not directly comparable
across different left-hand-side variables, so we rely on economic
reasoning and report the level specification as a robustness check.

Among log specifications, our preferred model is **(7)**: it has the
highest adjusted R-squared among specifications without bad controls,
and all its covariates have clear theoretical justification.

## 1b) Interpretation of the origin coefficients

```{r origin_coefs}
# Comment: We compute the mean French-origin wage as a reference point,
# then extract the origin coefficients from specification (7) and convert
# them to percentage effects and euro amounts.
mean_fr <- mean(df_q1$salhoraire[df_q1$origine == "France"], na.rm = TRUE)
cat("Mean hourly wage for French-origin individuals:", round(mean_fr, 2), "EUR\n")

coef_tab <- broom::tidy(reg7_log, conf.int = TRUE) %>%
  filter(grepl("^origine", term)) %>%
  mutate(
    pct_effect = (exp(estimate) - 1) * 100,
    euro_at_mean = pct_effect / 100 * mean_fr
  )
coef_tab %>%
  select(term, estimate, std.error, p.value, conf.low, conf.high,
         pct_effect, euro_at_mean)
```

In the preferred log specification (7), each coefficient on `origine` is
interpreted as a semi-elasticity: it gives the approximate percentage
wage gap between the corresponding origin group and individuals of
French origin, *holding age, sex, education, region, firm-specific
experience, parental CSP, marital status, self-reported health, urban
unit, and ZUS status fixed*. For coefficients small in absolute value
($|\hat\beta| < 0.10$), the approximation $\hat\beta \approx$
percentage effect holds; for larger coefficients, the exact formula
$(\exp(\hat\beta) - 1) \times 100$ is more accurate.

In all our specifications, the origin coefficients strongly depend on
the modality. Compared to individuals of French origin:

-   Some European-origin groups show a **positive** wage premium (e.g.
    Southern Europe, Northern Europe), which remains significant at the
    5% level in several specifications.
-   Other groups (notably Rest of Africa) show a **negative** wage gap,
    though its significance varies across specifications.
-   Most other modalities (Eastern Europe, Maghreb, Southeast Asia, RoW,
    Unknown) have point estimates between -5% and +2%, and are not
    individually significant at conventional thresholds.

However, non-significance for small subgroups (often fewer than 200
observations) should not be interpreted as evidence of no gap: it may
simply reflect low statistical power. Confidence intervals are more
informative than p-values in this context --- a 95% CI of [-12%, +4%]
for a group of 80 individuals tells us we cannot rule out a substantial
gap in either direction.

## 1c) Through which channels does origin affect wages?

It is likely that the effect of origin on wages is partly mediated
through other variables. To test this, we start from the Mincer baseline
(specification (6): origin + age + age² + sex + education + region) and
progressively add each candidate channel, tracking how the origin
coefficients move. The key channels we investigate are: parental social
background (`csp_pere_gp`), occupation (`csp_actif`), type of contract
(`contrat`), and type of working hours (`typehoraire`).

Note that adding `csp_actif`, `contrat`, and `typehoraire` is a form of
"bad control" analysis: these variables are themselves partly determined
by origin, so including them absorbs the *indirect* effect of origin
running through occupational sorting and contract segmentation. The
coefficient on origin after controlling for these channels captures only
the *residual direct* effect.

```{r mediation}
# Comment: We start from the Mincer baseline (demographics + education)
# so that education itself can also be studied as a channel.
m_base <- lm(logsalhoraire ~ origine + age + I(age^2) + homme + region,
             data = df_q1)

chan_specs <- list(
  "baseline (no educ)" = m_base,
  "+ education"        = update(m_base, . ~ . + education),
  "+ csp_pere_gp"      = update(m_base, . ~ . + education + csp_pere_gp),
  "+ csp_actif"        = update(m_base, . ~ . + education + csp_actif),
  "+ contrat"          = update(m_base, . ~ . + education + contrat),
  "+ typehoraire"      = update(m_base, . ~ . + education + typehoraire),
  "all channels"       = reg8_log
)

# Comment: Collect origin coefficients from each specification and plot
coef_evo <- map_dfr(chan_specs, ~ broom::tidy(.x), .id = "spec") %>%
  filter(grepl("^origine", term)) %>%
  mutate(spec = factor(spec, levels = names(chan_specs)))

ggplot(coef_evo,
       aes(x = spec, y = estimate, colour = term, group = term)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_line() +
  geom_point() +
  coord_flip() +
  labs(x = NULL,
       y = "Coefficient on log(hourly wage)",
       title = "Evolution of origin coefficients as channels are added") +
  theme_projet()
```

```{r mediation_table, results='asis'}
modelsummary(
  chan_specs,
  vcov = vcov_robust,
  stars = TRUE, output = "markdown",
  coef_omit = "^\\(Intercept\\)|^education|^region|^csp_pere_gp|^csp_actif|^contrat|^typehoraire|^matri|^sante_declaree|^tuu_r|^zus|^csp_mere_gp|^expe_specifique",
  title = "Sequential mediation analysis (log hourly wage)",
  gof_omit = "IC|Log|F|RMSE"
)
```

The plot and table make explicit which channels absorb the origin gap:

-   **Education**: adding education as a control reduces the negative
    coefficients for some origin groups (notably Rest of Africa), which
    is consistent with lower average educational attainment mediating
    part of the wage gap for these groups.

-   **Occupation (`csp_actif`)**: this is the single most powerful
    channel. Adding it typically halves the origin gap, consistent with
    origin groups being over-represented in lower-paying occupations.
    This does not mean origin has no effect on wages; it means the
    effect largely *operates through* occupational sorting.

-   **Contract type (`contrat`)**: controlling for contract type moves
    some coefficients further, pointing to a segmentation of labour
    contracts by origin (e.g. more fixed-term contracts for some groups).

-   **Parental CSP (`csp_pere_gp`)**: absorbs part of the gap for
    second-generation groups, consistent with intergenerational
    transmission of social status.

Importantly, "controlling for a mediator shrinks the coefficient" does
*not* mean the causal effect of origin is zero --- it means that part of
the total effect runs through that mediator.

## Question 2

### 2a) Why education is a problematic control

We can argue that education is partly explained by individual and family
choices that themselves depend on origin (e.g.cultural context, access
to information on education, resources, etc.). In other words, origin
impacts wages largely indirectly through its impact on education. By
adding education to the model, the estimated coefficient for origin only
captures the direct effect on wages, ignoring the indirect effect
through educational pathways that may be created by unobservable
factors. Consequently, we can expect the coefficient for origin to
become smaller in absolute value than the total effect, as it no longer
accounts for the potential disadvantages accumulated during the
individual's educational trajectory.

(An other phenomenon, althought that's not really our interest in this
question, is that the coefficient for education may be biased due to
omitted unobservable factors (such as innate talent or motivation) that
simultaneously influence both educational level and wages.)

### 2b) Choice of instruments

To address the main problem mentioned in (2a), we can use an
Instrumental Variable.

For a variable Z to be a valid instrument, it must satisfy two
conditions:
1)  Relevance: $Cov(\text{educ}, Z) \neq 0$. The instrument must be
    strongly correlated with education.
2)  Exogeneity: $Cov(\text{salhoraire}, Z) = 0$. Z should not be
    correlated with the error term; it should only impact wages
    indirectly through education. By intuition, csp_mere or csp_pere seem
    to be good choices: Relevance: Parental socio-professional category
    influences a lot the willingness and capacity to pursue long
    studies, often through social reproduction or different visions of
    job possibilities. Exogeneity: Parental CSP impacts wages primarily
    indirectly. Even if it shapes educational choices and cultural
    capital, an individual's current hourly wage is mostly determined by
    their own qualifications and career path, making any remaining
    direct influence from parents' status possibly negligible.

However, `nat_mere` and `nat_pere` are **not valid instruments**
here: `origine` is constructed from parental nationalities (see the data
appendix), so they are nearly collinear with the exogenous regressor
`origine` that we want to keep in the second stage. Using them as
instruments would violate the exclusion restriction.

To identify the causal effect of one exogeneous variable, we need one
valid instrument (one instrument for each exogeneous variable of our
model). Since we treat education as a categorical variable, each level
(represented by a dummy variable) is potentially endogenous. Therefore,
we need at least as many instruments as there are education levels
(minus the reference category). Using csp_mere as a categorical
instrument, for instance, provides enough instruments to identify the
effect of each education level (which is also the case for
nat_mere/pere)

However, the csp_mere and csp_pere variables have too many categories
(more than 30!). For the instruments to be robust, each category must
contain a significant number of individuals. To achieve this, we
group them by "Groupe Socioprofessionnel" (the 1st digit level)
according to the INSEE nomenclature. This grouping ensures we have
enough instruments to satisfy the order condition while maintaining
statistical power.

```{r}
df_q1 <- df_q1 %>%
  mutate(
    # Extract the first digit of parental CSP
    csp_pere_gp = substr(as.character(csp_pere), 1, 1),
    csp_mere_gp = substr(as.character(csp_mere), 1, 1)
  ) %>%
  mutate(
    # Convert to factors for the regression
    csp_pere_gp = as.factor(csp_pere_gp),
    csp_mere_gp = as.factor(csp_mere_gp)
  )

summary(df_q1$csp_mere_gp)
```

We have 8 instruments \> 5 exogenous variables (6 levels of education so
5 indicators/dummies in the model).

### 2c) 2SLS estimation

We estimate the 2SLS using the same controls as our preferred OLS
specification (7), so that the comparison between OLS and IV is
meaningful.

```{r iv_estimation}
# Instrument set 1: csp_pere_gp only
model_iv_1 <- ivreg(
  logsalhoraire ~ education + origine + age + I(age^2) + homme + region +
    expe_specifique + csp_mere_gp + matri + sante_declaree + tuu_r + zus |
    csp_pere_gp + origine + age + I(age^2) + homme + region +
    expe_specifique + csp_mere_gp + matri + sante_declaree + tuu_r + zus,
  data = df_q1)

# Instrument set 2: csp_pere_gp + csp_mere_gp
# Comment: csp_mere_gp moves from exogenous control to instrument here
model_iv_2 <- ivreg(
  logsalhoraire ~ education + origine + age + I(age^2) + homme + region +
    expe_specifique + matri + sante_declaree + tuu_r + zus |
    csp_pere_gp + csp_mere_gp + origine + age + I(age^2) + homme + region +
    expe_specifique + matri + sante_declaree + tuu_r + zus,
  data = df_q1)

summary(model_iv_1, diagnostics = TRUE)
summary(model_iv_2, diagnostics = TRUE)
```

```{r iv_comparison, results='asis'}
# Comment: Side-by-side display of OLS vs 2SLS for easy comparison
modelsummary(
  list("OLS (7)" = reg7_log,
       "2SLS (csp_pere)" = model_iv_1,
       "2SLS (csp_pere + csp_mere)" = model_iv_2),
  vcov = list(vcov_robust, vcov_robust, vcov_robust),
  stars = TRUE, output = "markdown",
  coef_omit = "^\\(Intercept\\)|^region|^matri|^sante_declaree|^tuu_r|^zus|^csp_pere_gp|^csp_mere_gp",
  title = "OLS vs 2SLS --- origin and education",
  gof_omit = "IC|Log|F|RMSE"
)
```

### 2d) Interpretation

Three points deserve comment:

1.  **Diagnostics.** The first-stage F-statistic on the excluded
    instruments should exceed the Stock--Yogo critical value (rule of
    thumb: F > 10) for the instruments to be considered strong. If the
    Sargan over-identification test rejects, at least one instrument is
    correlated with the wage residual --- most likely `csp_mere_gp`
    operating through professional networks --- and we should prefer the
    just-identified specification using only `csp_pere_gp`.

2.  **Education coefficient.** We expect the IV coefficient on education
    dummies to be *larger* in absolute value than OLS if the ability
    bias was attenuating the OLS estimate (Card-type argument), but also
    noisier due to the loss of efficiency from 2SLS.

3.  **Origin coefficient.** If the IV origin coefficients remain
    essentially unchanged compared to OLS after instrumenting education,
    it means the indirect-through-education channel is small for the
    compliers, and the bulk of the origin gap passes through other
    channels (occupation, contract type, parental networks) --- consistent
    with the mediation analysis of Question 1c. The wider standard errors
    are mechanical (2SLS is less efficient than OLS), so non-significance
    of individual origin modalities should not be over-interpreted; 95%
    confidence intervals are more informative than p-values here.
