---
title: "Project Econometrics - essais Aurélian 1"
author: "Aurélian"
date: "2026-04-08"
output: html_document
---

We install the general packages we need for the data manipulations
```{r}
if(!require(sjlabelled)) install.packages("sjlabelled")
if(!require(haven)) install.packages("haven")
if(!require(tidyverse)) install.packages("tidyverse")
if(!require(stargazer)) install.packages("stargazer")
if(!require(modelsummary)) install.packages("modelsummary")
if(!require(AER)) install.packages("AER")
library(haven) # to open a .dta file
library(tidyverse)
library(sjlabelled)
library(stargazer)
library(modelsummary)
library(dplyr)
library(AER)

```

We import the data
```{r}
df <- read_dta("DM_Subject_2_Data.dta")
```

We recreate, correctly, the "salhoraire" variable
```{r}
df$salhoraire = df$salmensuel / (df$heuretra * 4.33)
```

## Question 1

In the general case, the wage is registered for the first and the last quarters (T=1,2),
among six quarters (t=1,2,3,4,5,6).
Ici, we restrict us to the last quarter (ie, t=6).

We use filter to apply this restriction to our dataframe.
```{r}
df_q1 <- df %>% 
  filter(t==6)
```

### 1a) Estimation of the effect of some variables on wages

```{r}
summary(df_q1)

df_q1 <- df_q1 %>% 
  mutate(nat = as.numeric(nat),
         origine = as.numeric(origine),
         csp_actif = as.numeric(csp_actif))
```

**A First Regression**

To begin, we can try a first regression of salhoraire/logsalhoraire on variables that are commonly used in the literature to explain wages, that is to say: experience, age, gender, education and socio-professional category. We also include origin as it is the independent variable we are interested in.

```{r}

reg0_level <- lm(salhoraire ~ origine + age + homme + csp_actif+ education, df_q1)

reg0_log <- lm(logsalhoraire ~ origine + age + homme + csp_actif+ education, df_q1)
summary(reg0_level)
summary(reg0_log)

```
We obtain two models where most of the coefficients are statistically significant at the 5% threshold, and where the adjusted R-squared values are respectively 0.2625 and 0.3299 for the level and log-level models. We won't interpret them further for now, but we will give more details about why we include one variable or another, and we will then propose a few models and compare them.

**Some justification for the choice of variables**

***The independent variables of interest to be added to our model will be written in bold and italic.***

Minimal theoretical models:
log(salhoraire)= Beta_0 +Beta_origin*origin + u
ou
salhoraire= Beta_0 +Beta_origin*origin + u

One thing we want when estimating the effect of origin on individual's wages by linear regression is to find an unbiased estimator of the parameter associated to origin in such a regression.
For that to be verified, we need:
-1) the linearity in the parameter (we will specify our models to do so);
-2) the randomised sample (we have no hold over it but we may discuss it later);
-3) the non-perfect colinearity of the independent variables we are going to choose in our model (we will have to check it for each added variable);
-4) And then the nullity of the conditional expectation of the error term of the regression (E[epsilon|origin]=0), 

For (4) to be true, we will have to check if we have:
- a) the good specification of the functional form;
- b) no omitted variable;
(and (c) no inverse causality or (d) measurement errors - those two cases are more linked with the data construction and we will not really delay on them ).

Let's try to find an unbiased estimator of the effect of origin on salhoraire or logsalhoraire.


"annee" and "trimestre": it does not make sens to add them in the regression since we ignore the Panel dimension in this question

"dep" and "region": it seems economically plausible that the department/ region of the indivuduals influenced their wages, independently of their origin (do we need to test it by ANOVA or something since corr coeff is impossible???)
Moreover, these variables are certainly not perfectly colinear with the origin so it makes sens to include (one of) them, it will lower the omitted variable bias but not add perfect colinearity. The question is to know if it worth to add both. Indeed, since department are included in region we can believe that adding region will not add much information (let's try both and see which models are the best).
```{r}
#liste variables + labels
data.frame(Label = get_label(df))
```


"tuu"	tranche unite urbaine pour le logement, donnees 2010
"tuu_r"	tranche (regroupee) unite urbaine pour le logement, donnees 2010
"typuu"	type unite urbaine: R rural, U urbain, donnees 2010

These three variables deal with the urban categories in terms of density or number of housings. They might be correlated to both origin (cities are usually more metropolitan than countryside) and salhoraire (the cost and living and thus the wages are usually different whether you live in suburbs, city center or rural areas). Adding, at least, one of them could help to isolate the effect of origin on hourly wages. But, as before it might not be interesting adding them all in the model because they are redundant so it would not be clearly interpretable and it would add a lot of multicolinearity (if we suppose they are really correlated with origine) and thus increase the variance of beta_origine as well as those of Beta_tuu/tuu_r/typeuu without adding much information( well, the other thing we are interested in is to have a precise estimator (Beta_origine) that is to say with correct and small standard error ).

So we will try with ***"typuu" and/or "tuu_r"*** that are more easy to interpret.


***"zus"*** indicatrice d'appartenance à une zone urbaine sensible, can also be interesting to add for the same reasons as the previous variables (probably correlated with both origine (due to spatial segregation) and salhoraire(due to neighborhood effects, restricted job networks, or address-based discrimination)). Thus, omitting it would certainly create an omitted variable bias.

"taille_menage"	nombre de personnes dans le ménage
```{r}
cor(df_q1$taille_menage, df_q1$salhoraire, use = "complete.obs")
```
```{r}
# The correlation coefficient is very weak (~0.008). Since complete.obs removes rows with NAs during calculation, let's examine how the missing wage values are distributed across household sizes. This will help determine if this correlation coefficient is meaningful or if it suffers from selection bias.
table(df_q1$taille_menage)
sum(is.na(df_q1$salhoraire))

# Proportion of NAs (TRUE) vs. reported values (FALSE) by household size
prop.table(table(df_q1$taille_menage, is.na(df_q1$salhoraire)), margin = 1) * 100
```

There seems to be a non-response bias since the missing data rate varies by household size (though we should be careful as there are very few large households). In any case, the number of missing values is too high to trust the correlation coefficient. Can we follow economic logic instead ?

As we have no clear economic intuitions and we did not found information about links of wage and origin with number of people in a housing we will just try to add ***taille_menage*** in the model seeing if it add quality (R^2) and if is coefficient is significant but we will not fight for it to be in the model.

I HAVE NO OBVIOUS RELATIONS WITH TAILLE_MENAGE risque = corrélé à origine mais pas à salaire ????? mettre pas mettre

EST CE QUE DE MANIERE GENERALE CA FAIT SENS DE CALCULER COEF DE COR AVEC 5000 VALEURS MANQUANTES ET COMMENT ON FAIT POUR REG ? IMPUTATION ?

"csp_actif"
"csp_actif_int"
"csp_actif_detaille"

We include ***csp_actif*** as a control because it is strongly correlated with wages and is standard in the literature to isolate the effect of social status. We prefer the aggregate csp_actif over more detailed versions like csp_actif_int or csp_actif_detaille to avoid having too small sample sizes in specific sub-categories, which would lead to less robust estimates.

***"age"***
We will include age as a control variable because it serves as a proxy for accumulated professional experience. To not include age would bias the results, as any age differences between origin groups could be incorrectly attributed to their origin rather than their career stage.


"nat_mere"	nationalite de naissance de la mere
"nat_pere"	nationalite de naissance du pere
Those two variables gather (approximatively) the same information than origin, it would be useless (and dangerous -perfect colinearity) to add them.

"descimmi"	indicatrice d etre descendant d immigre
"immi"	indicatrice d etre immigre, ne etranger a l etranger

immi (1st Generation): These individuals may face specific barriers like non-recognized foreign diplomas, language difficulties, or a lack of local professional networks.

desc (2nd Generation): They were born and educated in France. If they still face wage gaps compared to the "Native" group, it suggests discrimination or social reproduction rather than a lack of integration or foreign diplomas.

As proposed in the appendix of the wording it could be interesting to add an interaction variable ***origin*imm***, we thus could see if the penalty of having a certain origin is the same for a 1st generation immigrant as it is for a 2nd generation descendant.

Moreover, we would not need to add desc as if immi = 0 since the information about being a descendant is already contained in the combination of origin != 01 and immi = 0.

"nat" nationnalité
Since our goal is more to test the effect of origin on hourly wages than the effect of nationnality (which would as well include some legal barrier to access some job in administrations for instance), we prefer not to include it.


"matri" statut matrimonial legal
"pacs" indicatrice d etre pacse

We include matri as a control variable because it is often correlated with both the individual’s origin (due to different cultural structures) and their hourly wage, as marital status can be a sign of stability associated with higher earnings.

Since matri has no missing values and is a more widespread indicator than pacs, which has 5,834 missing values, we prioritize adding ***matri*** to our model.

```{r}
# number of NA for matri and pacs
sum(is.na(df_q1$matri))
sum(is.na(df_q1$pacs))
```

***"homme"***
We include the gender indicator homme to control for the gender wage gap in the labor market.


"acteu"	statut sur le marche du travail au sens du BIT
"acteu_detail"	statut detaille sur le marche du travail au sens du BIT
"actif"	indicatrice d etre actif sur le marche du travail au sens du BIT
"actop"	indicatrice d etre actif occupe sur le marche du travail au sens du BIT
"chomeur"	indicatrice d etre chomeur sur le marche du travail au sens du BIT
"ancchom"	anciennete de chomage en mois
"ancinactif"	anciennete de l inactivite en mois
"halochomage"	halo autour du chomage
"informel"	indicatrice d exercice d une activite informelle durant sem. ref.
```{r}
# Création d'un tableau croisé entre le statut chômeur et la présence de NA dans le salaire
table(df_q1$chomeur, is.na(df_q1$salhoraire))
```
We observe that none of the 324 unemployed individuals have a reported wage (all are TRUE for is.na).Since the hourly wage is not recorded for the unemployed, there is no benefit in adding the indicator "chomeur" to the regression.

```{r}
# Tableau croisé entre le statut détaillé et la présence de salaire
# On utilise acteu_detail pour voir précisément qui ne répond pas
check_status <- table(df_q1$acteu_detail, is.na(df_q1$salhoraire), useNA = "always")

# On renomme les colonnes pour que ce soit plus clair
colnames(check_status) <- c("A un salaire (is.na = FALSE)", "Salaire Manquant (TRUE)", "NA Variable acteu_detail")
rownames(check_status) <- c("1-Employed", "3-Unemployed (Seeking)", "4-Other Unemployed", 
                            "5-Student", "6-Other Inactive", "NA Variable")

print(check_status)
```
Every unemployed individuals, students, and other inactives in the sample have no reported wage. Therefore, including these labor market status indicators in our regression is statistically irrelevant, as they provide no variation within the sub-sample of workers (we will have design matrix X which will have the column of acteu_detailed and intercept equal - so X'X not inversible).

We deduce that it is the same for "acteu", "actif" and "actop" so we cannot put any of them in our model.

```{r}
#nombre de NA par variable
sum(is.na(df_q1$ancchom))
sum(is.na(df_q1$ancinactif))
sum(is.na(df_q1$informel))
# On regarde le nombre de NA pour ancchom,ancinactif et informel  uniquement pour ceux qui ont un salaire
sum(is.na(df_q1$ancchom[is.na(df_q1$salhoraire) == FALSE]))
sum(is.na(df_q1$ancinactif[is.na(df_q1$salhoraire) == FALSE]))
sum(is.na(df_q1$informel[is.na(df_q1$salhoraire) == FALSE]))
```
Since ancchom, informel and ancinactif have more than 5,000 NA out of 10,000 observations (which are for most of them different from the NA of salhoraire) and are not central to answering Question 1, we prioritize maintaining a larger and more representative sample and do not add them.


***"education"***	diplome le plus eleve obtenu (6 postes)
"education_detail1"	diplome le plus eleve obtenu (16 postes)
"education_detail2"	diplome le plus eleve obtenu (11 postes)

Education is surely correlated with hourly wage, and adding it allows us to isolate the effect of origin from educational attainment. For clarity, and to be sure to have enough individuals in each category for a robust estimation, we choose the education variable (6 categories) over the more detailed versions.



***expe_specifique***	anciennete dans l entreprise (ou fonction publique) en mois
```{r}
sum(is.na(df_q1$expe_specifique[is.na(df_q1$salhoraire) == FALSE]))
```
It is also very relevant to include expe_specifique because it controls for the increase in the wage linked with firm-specific skills, allowing us to compare individuals of different origins at equal levels of stability in terms of staying in a company.


***plract***	indicatrice d exercer plusieurs activites professionnelles distinctes
We include plract (multiple job-holding) as a control variable. Holding more than one job may indicate financial necessity and might be correlated to having low wages. Since this propensity might also differ by origin, controlling for it can be interesting.
```{r}
sum(is.na(df_q1$plract[is.na(df_q1$salhoraire) == FALSE]))
```

***csp_mere***	CSP de la mere (niveau detaille 2 chiffres)
***csp_pere***	CSP du pere (niveau detaille 2 chiffres)

We include both csp_pere and csp_mere to control for social background, ensuring that the estimated effect of origin is not biased by the some social reproduction of professional status from the parents. BOTH ???? - agregate them ????

***sante_declaree***	appreciation personnelle de son etat de sante, NB a t=1,6 uniquement

We include sante_declaree as it is fully reported at t=6 and serves as a proxy for individual productivity, which directly impacts hourly wages.

We acknowledge potential endogeneity issues due to reverse causality in variables such as sante_declaree. For instance, while health may impact productivity, higher earnings also provide better access to healthcare so sante_declare could be correlated with the error term.


"heuretra"	nombre d heures travaillees en moyenne par semaine dans l emploi principal, decl
"heuretra_calculees"	nombre d heures travaillees en moyenne par semaine, calcul nb jour x 7
"diff_heuretra"	difference heuretra - heuretra_calculees
"heuretra_tranche"	nombre d heures travaillees en moyenne par semaine dans l emploi principal, en t
"typehoraire"	type d horaires effectues dans l emploi principal
"heuretrasouhait"	nombre heure de travail souhaite dans l ideal, avec variation associee de revenu
"compenheuresup"	nature de compensation auxquelles donnent droit les heures supplementaires
"jourtra"	nombre de jours habituellement travailles par semaine

We exclude heuretra, jourtra and salmensuel since salhoraire is already built on wage/time worked, creating a model would lose part of its interest. However, we can try to include ***typehoraire*** and ***compenheuressup*** to control for job-specific characteristics and compensation schemes—such that naturally influence the hourly wage. (PHRASE POURRIE TROUVER MEILLEURS ARGUMENTS)

Des formes fonctionnelles particulières:

It is likely that age and experience have non-linear relationships with hourly wages, as their marginal impact typically increases during the early career before eventually diminishing. We will test whether a quadratic specification is appropriate to capture these diminishing returns.


```{r}
plot(df_q1$age, df_q1$salhoraire,
     xlab = "Age", ylab = "Hourly wage",
     main = "Relation between Age and Hourly Wage",
     pch = 19, col = "blue", cex = 0.5)
```
```{r}
plot(df_q1$age, df_q1$logsalhoraire,
     xlab = "Age", ylab = "log of Hourly wage",
     main = "Relation between Age and log of Hourly Wage",
     pch = 19, col = "blue", cex = 0.5)
```

The concave, bell-shaped pattern of the scatter plot  suggests testing a quadratic specification for age in the Level model and the Log Level one.

```{r}
summary(lm(salhoraire ~ age + I(age^2), data=df_q1))
```
Le coefficient associé à age^2 n'est pas significatif à aucun seuil ----- donc quadratique à nuancer on ne sais pas ??
```{r}
plot(df_q1$expe_specifique, df_q1$salhoraire,
     xlab = "experience", ylab = "Hourly wage",
     main = "Relation between Experience and Hourly Wage",
     pch = 19, col = "blue", cex = 0.5)
```
```{r}
plot(df_q1$expe_specifique, df_q1$logsalhoraire,
     xlab = "experience", ylab = "Log Hourly wage",
     main = "Relation between Experience and log Hourly Wage",
     pch = 19, col = "blue", cex = 0.5)

```


```{r}
#courbe de régression simple non paramétrique sur des sous ensembles de points (~LOESS)
library(ggplot2)
ggplot(df_q1, aes(age, logsalhoraire)) + 
  geom_smooth(method = "lm", formula = y ~ poly(x, 2))#???revoir
```

The quadratic form makes less sens for experience in the level and log level models.


```{r}
names(df_q1)
```
```{r}
sum(is.na(df_q1$csp_actif))
```


**Tests de régression:**



```{r}
# Converting numerical codes to factors for correct interpretation
df_q1 <- df_q1 %>% 
  mutate(
    # Variables of interest 
    origine = as.factor(origine),
    immi = as.factor(immi),
    
    # Categorical controls (Dummies)
    csp_actif = as.factor(csp_actif),
    education = as.factor(education),
    homme = as.factor(homme),
    matri = as.factor(matri),
    zus = as.factor(zus),
    tuu_r = as.factor(tuu_r),
    typuu = as.factor(typuu),
    sante_declaree = as.factor(sante_declaree),
    csp_pere = as.factor(csp_pere),
    csp_mere = as.factor(csp_mere),
    typehoraire = as.factor(typehoraire),
    compenheuresup = as.factor(compenheuresup)
  )

# INCREMENTAL REGRESSION MODELS

# 1. Baseline Model: Standard predictors in the literature
reg0_level <- lm(salhoraire ~ origine + age + homme + csp_actif + education, data = df_q1)

# 2. Migration Interaction: Testing if the origin effect differs between immigrants and descendants of immigrants
reg1_level <- lm(salhoraire ~ origine * immi + age + homme + csp_actif + education, data = df_q1)

# 3. Quadratic Age: Capturing non-linear diminishing returns to experience
reg2_level <- lm(salhoraire ~ origine * immi + age + I(age^2) + homme + csp_actif + education, data = df_q1)

# 4. Firm-Specific Experience: Controlling for tenure-based wage premiums
reg3_level <- lm(salhoraire ~ origine * immi + age + I(age^2) + homme + csp_actif + education + expe_specifique, data = df_q1)

# 5. Social Background: Adding parental CSP to isolate origin from social reproduction
reg4_level <- lm(salhoraire ~ origine * immi + age + I(age^2) + homme + csp_actif + education + csp_pere + csp_mere, data = df_q1)

# 6. Health Capital: Proxy for individual productivity (reported at t=6)
reg5_level <- lm(salhoraire ~ origine * immi + age + I(age^2) + homme + csp_actif + education + sante_declaree, data = df_q1)

# 7. Household Structure: Controlling for marital status and household size
reg6_level <- lm(salhoraire ~ origine * immi + age + I(age^2) + homme + csp_actif + education + matri + taille_menage, data = df_q1)

# 8. Spatial Segregation: Controlling for neighborhood effects (ZUS) and urban density (TUU)
reg7_level <- lm(salhoraire ~ origine * immi + age + I(age^2) + homme + csp_actif + education + zus + tuu_r, data = df_q1)

# 9. Working Conditions: Including compensating differentials (schedules, overtime, plract)
reg8_level <- lm(salhoraire ~ origine * immi + age + I(age^2) + homme + csp_actif + education + typehoraire + compenheuresup + plract, data = df_q1)

# 10. Comprehensive Synthesis: Combining the most economically significant controls
reg9_level <- lm(salhoraire ~ origine * immi + age + I(age^2) + homme + csp_actif + education + 
                   expe_specifique + csp_pere + sante_declaree + zus + tuu_r + typehoraire, data = df_q1)
summary(reg0_level)
```

```{r}
# --- LOG-LEVEL INCREMENTAL MODELS ---

# 1. Baseline Model
reg0_log <- lm(logsalhoraire ~ origine + age + homme + csp_actif + education, data = df_q1)

# 2. Migration Interaction (The "Interest" variables)
reg1_log <- lm(logsalhoraire ~ origine * immi + age + homme + csp_actif + education, data = df_q1)

# 3. Quadratic Age (Capturing the concave earnings profile)
reg2_log <- lm(logsalhoraire ~ origine * immi + age + I(age^2) + homme + csp_actif + education, data = df_q1)

# 4. Firm-Specific Experience
reg3_log <- lm(logsalhoraire ~ origine * immi + age + I(age^2) + homme + csp_actif + education + expe_specifique, data = df_q1)

# 5. Social Background (Parental CSP)
reg4_log <- lm(logsalhoraire ~ origine * immi + age + I(age^2) + homme + csp_actif + education + csp_pere + csp_mere, data = df_q1)

# 6. Health Capital
reg5_log <- lm(logsalhoraire ~ origine * immi + age + I(age^2) + homme + csp_actif + education + sante_declaree, data = df_q1)

# 7. Household Structure
reg6_log <- lm(logsalhoraire ~ origine * immi + age + I(age^2) + homme + csp_actif + education + matri + taille_menage, data = df_q1)

# 8. Spatial Segregation (ZUS & Urban Density)
reg7_log <- lm(logsalhoraire ~ origine * immi + age + I(age^2) + homme + csp_actif + education + zus + tuu_r, data = df_q1)

# 9. Working Conditions (including multiple job-holding 'plract')
reg8_log <- lm(logsalhoraire ~ origine * immi + age + I(age^2) + homme + csp_actif + education + typehoraire + compenheuresup + plract, data = df_q1)

# 10. Synthesis (arbitrage sur redondance) Model
reg9_log <- lm(logsalhoraire ~ origine * immi + age + I(age^2) + homme + csp_actif + education + 
                 expe_specifique + csp_pere + sante_declaree + zus + tuu_r + typehoraire, data = df_q1)
```



```{r}


# Compare the two using AIC 
# Note: To compare AIC of log vs level, you must adjust the log-AIC manually, ?!!!!!!!!!!!!!!!!!
# but usually, we choose based on the distribution of residuals.?????? revoir
```
```{r}
# Create a table with the coefficients and indicator of each model
models <- list(
  # Level models
  "M0_Lev" = reg0_level, "M1_Lev" = reg1_level, "M2_Lev" = reg2_level, "M3_Lev" = reg3_level,
  "M4_Lev" = reg4_level, "M5_Lev" = reg5_level, "M6_Lev" = reg6_level, "M7_Lev" = reg7_level,
  "M8_Lev" = reg8_level, "M9_Lev" = reg9_level,
  # Log models
  "M0_Log" = reg0_log, "M1_Log" = reg1_log, "M2_Log" = reg2_log, "M3_Log" = reg3_log,
  "M4_Log" = reg4_log, "M5_Log" = reg5_log, "M6_Log" = reg6_log, "M7_Log" = reg7_log,
  "M8_Log" = reg8_log, "M9_Log" = reg9_log
)

modelsummary(models, 
             stars = TRUE, 
             title = "Résultats des régressions")
```

FINALEMENT EXPLIQUER LES RAISONS (R**2) et éco qui font qu'on choisit parmi log-niveau puis choisir parmi aic R**2 significativité des coefs....



Du point de vue R**2 et des AIC/BIC M3 Log est le mieux -- petit problème les coefficients associés aux origines ne sont significatifs statistiquement DANS AUCUN MODELE ---POSSIBLE DE CREER UNE VARIABLE BINAIRE ORIGINE IMIGRE OU C'EST PAS CE QUI VEULENT ?


The best model, regarding both its explainibility (through the R^2) and its parsimony (through 
the AIC and the number of variables) seems to be the LOG_3 model. Indeed, the most efficient
strategy consists in selecting only some classic variables used in the econometrics of the wages,
ie sex, age, education and experience. 

However, the log-level specification is better than the level-level one. For example, the effect of sex on the wage is probably quite similar accross the distribution of the wages : whatever the professional sector, men are generally slightly better paid than women. Therefore, being a man leads to a greater average wage, following a certain percentage of difference between sex, and not to a fixed amount of money in addition. That is why the effect of sex on wages is logarithmic : the higher the wage, the higher the effect of sex on wage in level. 

However, the specification is not perfectly log-linear, as the age has mainly a quadratic effect. This can be easily explained by the life cycle theory : the wage increases with age until a pic, and then decreases until retirement and even fastly until death. 





#### 1b)

In all our models, the coefficients of the "origine" variable strongly depend on the modality. In other words, compared to being from France, being from another country can have an effect on wage, but this effect is either positive or negative according the country that is considered. 

For example, being from Southern Europe (origine = 4) seems to have, in each model, a positive effect on wage compared to being from France (origine = 1). The effect is opposed for the fact of being from Eastern Europe (origine = 5). 

Nonetheless, almost all the modalities of this variable have a non-significant effect. The fact of being from Southern Europe is, indeed, the only modality that has a significant effect, at the 5% threshold and mainly for the level-level specificated model. On the contrary, the effect of being from Africa or from Middle East is never significant, whatever the model is. 



#### 1c)

It is likely that the effect of the origin on wages is mediated through other variables. Let construct some models to test that hypothesis.

```{r}
reg0_test_level <- lm(salhoraire ~ origine, df_q1)
reg1_test_level <- lm(salhoraire ~ origine + education, df_q1)
reg2_test_level <- lm(salhoraire ~ origine + contrat, df_q1)
reg3_test_level <- lm(salhoraire ~ origine + csp_actif, df_q1)

reg0_test_log <- lm(logsalhoraire ~ origine, df_q1)
reg1_test_log <- lm(logsalhoraire ~ origine + education, df_q1)
reg2_test_log <- lm(logsalhoraire ~ origine + contrat, df_q1)
reg3_test_log <- lm(logsalhoraire ~ origine + csp_actif, df_q1)

summary(reg0_test_level)
summary(reg1_test_level)
summary(reg2_test_level)
summary(reg3_test_level)

summary(reg0_test_log)
summary(reg1_test_log)
summary(reg2_test_log)
summary(reg3_test_log)



```
```{r}
models_test <- list(
  # Level models
  "M0_Test_Lev" = reg0_test_level, "M1_Test_Lev" = reg1_test_level, "M2_Test_Lev" = reg2_test_level, "M3_Test_Lev" = reg3_test_level,
  # Log models
  "M0_Test_Log" = reg0_test_log, "M1_Test_Log" = reg1_test_log, "M2_Test_Log" = reg2_test_log, "M3_Test_Log" = reg3_test_log)

modelsummary(models_test, 
             stars = TRUE, 
             title = "Résultats des régressions")

```


The comparison between these four models is instructive. Indeed, one can see a concrete mediation of the effect of the origin through other variables, which is not univoque.

For example, in the first level-level model, we see that only 7 and 99 (respectively rest of Africa and unknown) have a significant impact, negative in that case. When we control by education, the effect is lower for rest of Africa and is even no longer significant for unknown. One can assume that people from Africa or whose nationality is unknown have more often a low education than other people, which means that the effect of their origin on their wage is partly mediated through their low education. 


But controlling by another variable can also make the effect some modalities significant. Typically, when one controls by the type of contract, the modality 3 (Northern Europe) begins to have a significant and positive impact on wages. This probably means that, among a group of people having the same type of contract, people from Northern Europe have a better wage, in average.

Finally, controlling by the occupation (by the "PCS" in particular) seems to be quite relevant, as the effect of origin is made non-significant that way, no matter what modality is considered. This was previsible, given that the places of people having some origins in the labour market are often different from people coming from France. For example, people from Africa are more often employed in the industry, at some subalterned positions, etc. 



## Question 2
#### 2a)

We can argue that education is partly explained by individual and family choices that themselves depend on origin (e.g.cultural context, access to information on education, resources, etc.). In other words, origin impacts wages largely indirectly through its impact on education. By adding education to the model, the estimated coefficient for origin only captures the direct effect on wages, ignoring the indirect effect through educational pathways that may be created by unobservable factors. Consequently, we can expect the coefficient for origin to become smaller in absolute value than the total effect, as it no longer accounts for the potential disadvantages accumulated during the individual's educational trajectory.

(An other phenomenon, althought that's not really our interest in this question, is that the coefficient for education may be biased due to omitted unobservable factors (such as innate talent or motivation) that simultaneously influence both educational level and wages.)

#### 2b)
To avoid the main problem mentioned in (2a), we can add an instrument for education level: 
For an additionnal variable Z to be an instrument we need:
- Pertienence : Cov(educ, Z) different from 0 and even as strongly correlated as possible;
- Exogeneity: Cov(salhoraire, Z)=0, Z should not be correlated to salhoraire (just indirectly via educ);

By intuition, csp_mere or csp_pere seem to bee good choices:
The socioprofessional category of the parents influenced a lot the choices and willingness to pursue long or short studies by reproduction of choices of parents or by the vision on job possibilities you have for instance.
The socioprofessional category might influence a bit the wage but really indirectly since even if it influenced the educational choices or even the cultural capital of the children their current job and wage is for from being only influenced by these two things.

To address the main problem mentioned in (2a), we can use an Instrumental Variable. 

For a variable Z to be a valid instrument, it must satisfy two conditions:

1) Relevance: $Cov(\text{educ}, Z) \neq 0$. The instrument must be strongly correlated with education.
2) Exogeneity: $Cov(\text{salhoraire}, Z) = 0$. Z should not be correlated with the error term; it should only impact wages indirectly through education.By intuition, csp_mere or csp_pere seem to be good choices:
Relevance: Parental socio-professional category influences a lot the willingness and capacity to pursue long studies, often through social reproduction or different visions of job possibilities.
Exogeneity: Parental CSP impacts wages primarily indirectly. Even if it shapes educational choices and cultural capital, an individual's current hourly wage is mostly determined by their own qualifications and career path, making any remaining direct influence from parents' status possibly negligible.

In the same vein, we can think to nat_mere or nat_pere:
Relevance: Cultural background and access to specific scholarships or administrative support due to a parent's nationality can influence the child's educational level.

Exogeneity: Aside from its impact on education , we can believe that a parent's nationality does not influence the innate talent or productivity of their children at work, and therefore does not directly influence their wage. There might remains a potential small residual discrimination based on foreign name, etc. so these instrument might be a little less valid than csp_mere or csp_pere. 


OTHER IDEAS OF RESIDUALS??

To identify the causal effect of one exogeneous variable, we need one valid instrument (one instrument for each exogeneous variable of our model).
Since we treat education as a categorical variable, each level (represented by a dummy variable) is potentially endogenous. Therefore, we need at least as many instruments as there are education levels (minus the reference category). Using csp_mere as a categorical instrument, for instance, provides enough instruments to identify the effect of each education level (which is also the case for nat_mere/pere)

However, the csp_mere and csp_pere variables have too many categories (more than 30!). For the instruments to be robust, each category must contain a significant number of individuals. To achieve this, we will group them by "Groupe Socioprofessionnel" (the 1st digit level) according to the INSEE nomenclature. This grouping ensures we have enough instruments to satisfy the order condition while maintaining statistical power.


```{r} 
# concatenation according to INSEE categories "Groupes Socioprofessionnel" (we just group by the first digit of each code)


df_q1 <- df_q1 %>%
  mutate(
    # On prend le premier chiffre de la CSP du père et de la mère
    csp_pere_gp = substr(as.character(csp_pere), 1, 1),
    csp_mere_gp = substr(as.character(csp_mere), 1, 1)
  ) %>%
  mutate(
    # On les transforme en facteurs pour la régression
    csp_pere_gp = as.factor(csp_pere_gp),
    csp_mere_gp = as.factor(csp_mere_gp)
  )

summary(df_q1$csp_mere_gp)
```
We have 8 instruments > 5 exogenous variables (6 levels of education so 5 indicators/dummies in the model).


###2c) 
We are going to use the 2SLS to estimate these model (trough function ivreg).
```{r}
library(AER)

# 2SLS Model 
model_iv_1 <- ivreg(logsalhoraire ~ education + origine |  csp_mere_gp + origine, data = df_q1)
model_iv_1 <- ivreg(logsalhoraire ~ education + origine |  csp_mere_gp + csp_pere_gp + origine, data = df_q1)
model_iv_2 <- ivreg(logsalhoraire ~ education + origine |  nat_pere + origine, data = df_q1)
# Pour voir les résultats et les tests de diagnostic (Sargan, Weak instruments) ???? nécessaire ??? revoir
summary(model_iv_1, diagnostics = TRUE)
```
#INTErprétation à finir  + voir si 2C) veulent meme modele que Q1 ou non
```{r}
#2SLS sur même variables de contrôles que Reg3_log, le meilleur modèle de la question 1a)
reg3_log <- lm(logsalhoraire ~ origine * immi + age + I(age^2) + homme + csp_actif + education + expe_specifique, data = df_q1)
model_iv_222 <- ivreg(logsalhoraire ~ origine * immi + age + I(age^2) + homme + csp_actif + education + expe_specifique |  csp_mere_gp + origine * immi + age + I(age^2) + homme + csp_actif + expe_specifique , data = df_q1)
summary(model_iv_222, diagnostics = TRUE)
```
Arghhhhhhhhhhhhh tjrs rien de significatif, quelle interprétation ? --- Once we control for the effect of education, origin has no statistically significant effect on the wage ( compared to the previous OLS models (above all the smaller ones) where origine7 and sometimes origine3 or 4 where significant that is no longer the case here ) ???

