---
title: "Project Econometrics - essais Aurûˋlian 1"
author: "Aurûˋlian"
date: "2026-04-08"
output: html_document
---

We install the general packages we need for the data manipulations
```{r}
if(!require(sjlabelled)) install.packages("sjlabelled")
if(!require(haven)) install.packages("haven")
if(!require(tidyverse)) install.packages("tidyverse")
if(!require(stargazer)) install.packages("stargazer")
if(!require(modelsummary)) install.packages("modelsummary")
library(haven) # to open a .dta file
library(tidyverse)
library(sjlabelled)
library(stargazer)
library(modelsummary)

```

We import the data
```{r}
df <- read_dta("DM_Subject_2_Data.dta")
```

We recreate, correctly, the "salhoraire" variable
```{r}
data$salhoraire = data$salmensuel / (data$heuretra * 4.33)
```

Question 1

In the general case, the wage is registered for the first and the last quarters (T=1,2),
among six quarters (t=1,2,3,4,5,6).
Ici, we restrict us to the last quarter (ie, t=6).

We use filter to apply this restriction to our dataframe.
```{r}
df_q1 <- df %>% 
  filter(t==6)
```

1a) Estimation of the effect of some variables on wages

```{r}
summary(df_q1)

df_q1 <- df_q1 %>% 
  mutate(nat = as.numeric(nat),
```


```{r}
```


```{r}
origine = as.numeric(origine),
         csp_actif = as.numeric(csp_actif))

reg1 <- lm(salhoraire ~ nat + origine + age + homme + csp_actif, df_q1)
summary(reg1)

```

Minimal theoretical models:
log(salhoraire)= Beta_0 +Beta_origin*origin + u
ou
salhoraire= Beta_0 +Beta_origin*origin + u

One thing we want when estimating the effect of origin on individual's wages by linear regression is to find an unbiased estimator of the parameter associated to origin in such a regression.
For that to be verified, we need:
-1) the linearity in the parameter (we will specify our models to do so);
-2) the randomised sample (we have no hold over it but we may discuss it later);
-3) the non-perfect colinearity of the independent variables we are going to choose in our model (we will have to check it for each added variable);
-4) And then the nullity of the conditional expectation of the error term of the regression (E[epsilon|origin]=0), 

For (4) to be true, we will have to check if we have:
- a) the good specification of the functional form;
- b) no omitted variable;
(and (c) no inverse causality or (d) measurement errors - those two cases are more linked with the data construction and we will not really delay on them ).

Let's try to find an unbiased estimator of the effect of origin on salhoraire or logsalhoraire.


"annee" and "trimestre": it does not make sens to add them in the regression since we ignore the Panel dimension in this question

"dep" and "region": it seems economically plausible that the department/ region of the indivuduals influenced their wages, independently of their origin (do we need to test it by ANOVA or something since corr coeff is impossible???)
Moreover, these variables are certainly not perfectly colinear with the origin so it makes sens to include (one of) them, it will lower the omitted variable bias but not add perfect colinearity. The question is to know if it worth to add both. Indeed, since department are included in region we can believe that adding region will not add much information (let's try both and see which models are the best).
```{r}
#liste variables + labels
data.frame(Label = get_label(df))
```


"tuu"	tranche unite urbaine pour le logement, donnees 2010
"tuu_r"	tranche (regroupee) unite urbaine pour le logement, donnees 2010
"typuu"	type unite urbaine: R rural, U urbain, donnees 2010

These three variables deal with the urban categories in terms of density or number of housings. They might be correlated to both origin (cities are usually more metropolitan than countryside) and salhoraire (the cost and living and thus the wages are usually different whether you live in suburbs, city center or rural areas). Adding, at least, one of them could help to isolate the effect of origin on hourly wages. But, as before it might not be interesting adding them all in the model because they are redundant so it would not be clearly interpretable and it would add a lot of multicolinearity (if we suppose they are really correlated with origine) and thus increase the variance of beta_origine as well as those of Beta_tuu/tuu_r/typeuu without adding much information( well, the other thing we are interested in is to have a precise estimator (Beta_origine) that is to say with correct and small standard error ).

So we will try with "typuu" and/or "tuu_r" that are more easy to interpret.


"zus" indicatrice d'appartenance û  une zone urbaine sensible, can also be interesting to add for the same reasons as the previous variables (probably correlated with both origine (due to spatial segregation) and salhoraire(due to neighborhood effects, restricted job networks, or address-based discrimination)). Thus, omitting it would certainly create an omitted variable bias.

"taille_menage"	nombre de personnes dans le mûˋnage
```{r}
cor(df_q1$taille_menage, df_q1$salhoraire, use = "complete.obs")
```
```{r}
# The correlation coefficient is very weak (~0.008). Since complete.obs removes rows with NAs during calculation, let's examine how the missing wage values are distributed across household sizes. This will help determine if this correlation coefficient is meaningful or if it suffers from selection bias.
table(df_q1$taille_menage)
sum(is.na(df_q1$salhoraire))

# Proportion of NAs (TRUE) vs. reported values (FALSE) by household size
prop.table(table(df_q1$taille_menage, is.na(df_q1$salhoraire)), margin = 1) * 100
```

There seems to be a non-response bias since the missing data rate varies by household size (though we should be careful as there are very few large households). In any case, the number of missing values is too high to trust the correlation coefficient. Can we follow economic logic instead ?

As we have no clear economic intuitions and we did not found information about links of wage and origin with number of people in a housing we will just try to add taille_menage in the model seeing if it add quality (R^2) and if is coefficient is significant but we will not fight for it to be in the model.

I HAVE NO OBVIOUS RELATIONS WITH TAILLE_MENAGE risque = corrûˋlûˋ û  origine mais pas û  salaire ????? mettre pas mettre

EST CE QUE DE MANIERE GENERALE CA FAIT SENS DE CALCULER COEF DE COR AVEC 5000 VALEURS MANQUANTES ET COMMENT ON FAIT POUR REG ? IMPUTATION ?



"age"
We will include age as a control variable because it serves as a proxy for accumulated professional experience. To not include age would bias the results, as any age differences between origin groups could be incorrectly attributed to their origin rather than their career stage.


"nat_mere"	nationalite de naissance de la mere
"nat_pere"	nationalite de naissance du pere
Those two variables gather (approximatively) the same information than origin, it would be useless (and dangerous -perfect colinearity) to add them.

"descimmi"	indicatrice d etre descendant d immigre
"immi"	indicatrice d etre immigre, ne etranger a l etranger

immi (1st Generation): These individuals may face specific barriers like non-recognized foreign diplomas, language difficulties, or a lack of local professional networks.

desc (2nd Generation): They were born and educated in France. If they still face wage gaps compared to the "Native" group, it suggests discrimination or social reproduction rather than a lack of integration or foreign diplomas.

As proposed in the appendix of the wording it could be interesting to add an interaction variable origin*immi, we thus could see if the penalty of having a certain origin is the same for a 1st generation immigrant as it is for a 2nd generation descendant.

Moreover, we would not need to add desc as if immi = 0 since the information about being a descendant is already contained in the combination of origin != 01 and immi = 0.

"nat" nationnalitûˋ
Since our goal is more to test the effect of origin on hourly wages than the effect of nationnality (which would as well include some legal barrier to access some job in administrations for instance), we prefer not to include it.


"matri" statut matrimonial legal
"pacs" indicatrice d etre pacse

We include matri as a control variable because it is often correlated with both the individualãs origin (due to different cultural structures) and their hourly wage, as marital status can be a sign of stability associated with higher earnings.

Since matri has no missing values and is a more widespread indicator than pacs, which has 5,834 missing values, we prioritize adding matri to our model.

```{r}
# number of NA for matri and pacs
sum(is.na(df_q1$matri))
sum(is.na(df_q1$pacs))
```

"homme"
We include the gender indicator homme to control for the gender wage gap in the labor market.


"acteu"	statut sur le marche du travail au sens du BIT
"acteu_detail"	statut detaille sur le marche du travail au sens du BIT
"actif"	indicatrice d etre actif sur le marche du travail au sens du BIT
"actop"	indicatrice d etre actif occupe sur le marche du travail au sens du BIT
"chomeur"	indicatrice d etre chomeur sur le marche du travail au sens du BIT
"ancchom"	anciennete de chomage en mois
"ancinactif"	anciennete de l inactivite en mois
"halochomage"	halo autour du chomage
"informel"	indicatrice d exercice d une activite informelle durant sem. ref.
```{r}
# Crûˋation d'un tableau croisûˋ entre le statut chûÇmeur et la prûˋsence de NA dans le salaire
table(df_q1$chomeur, is.na(df_q1$salhoraire))
```
We observe that none of the 324 unemployed individuals have a reported wage (all are TRUE for is.na).Since the hourly wage is not recorded for the unemployed, there is no benefit in adding the indicator "chomeur" to the regression.

```{r}
# Tableau croisûˋ entre le statut dûˋtaillûˋ et la prûˋsence de salaire
# On utilise acteu_detail pour voir prûˋcisûˋment qui ne rûˋpond pas
check_status <- table(df_q1$acteu_detail, is.na(df_q1$salhoraire), useNA = "always")

# On renomme les colonnes pour que ce soit plus clair
colnames(check_status) <- c("A un salaire (is.na = FALSE)", "Salaire Manquant (TRUE)", "NA Variable acteu_detail")
rownames(check_status) <- c("1-Employed", "3-Unemployed (Seeking)", "4-Other Unemployed", 
                            "5-Student", "6-Other Inactive", "NA Variable")

print(check_status)
```
Every unemployed individuals, students, and other inactives in the sample have no reported wage. Therefore, including these labor market status indicators in our regression is statistically irrelevant, as they provide no variation within the sub-sample of workers (we will have design matrix X which will have the column of acteu_detailed and intercept equal - so X'X not inversible).

We deduce that it is the same for "acteu", "actif" and "actop" so we cannot put any of them in our model.

```{r}
#nombre de NA par variable
sum(is.na(df_q1$ancchom))
sum(is.na(df_q1$ancinactif))
sum(is.na(df_q1$informel))
# On regarde le nombre de NA pour ancchom,ancinactif et informel  uniquement pour ceux qui ont un salaire
sum(is.na(df_q1$ancchom[is.na(df_q1$salhoraire) == FALSE]))
sum(is.na(df_q1$ancinactif[is.na(df_q1$salhoraire) == FALSE]))
sum(is.na(df_q1$informel[is.na(df_q1$salhoraire) == FALSE]))
```
Since ancchom, informel and ancinactif have more than 5,000 NA out of 10,000 observations (which are for most of them different from the NA of salhoraire) and are not central to answering Question 1, we prioritize maintaining a larger and more representative sample and do not add them.


"education"	diplome le plus eleve obtenu (6 postes)
"education_detail1"	diplome le plus eleve obtenu (16 postes)
"education_detail2"	diplome le plus eleve obtenu (11 postes)

Education is surely correlated with hourly wage, and adding it allows us to isolate the effect of origin from educational attainment. For clarity, and to be sure to have enough individuals in each category for a robust estimation, we choose the education variable (6 categories) over the more detailed versions.



expe_specifique	anciennete dans l entreprise (ou fonction publique) en mois
```{r}
sum(is.na(df_q1$expe_specifique[is.na(df_q1$salhoraire) == FALSE]))
```
It is also very relevant to include expe_specifique because it controls for the increase in the wage linked with firm-specific skills, allowing us to compare individuals of different origins at equal levels of stability in terms of staying in a company.

csp_mere	CSP de la mere (niveau detaille 2 chiffres)
csp_pere	CSP du pere (niveau detaille 2 chiffres)

We include both csp_pere and csp_mere to control for social background, ensuring that the estimated effect of origin is not biased by the some social reproduction of professional status from the parents. BOTH ???? - agregate them ????

sante_declaree	appreciation personnelle de son etat de sante, NB a t=1,6 uniquement

We include sante_declaree as it is fully reported at t=6 and serves as a proxy for individual productivity, which directly impacts hourly wages.


"heuretra"	nombre d heures travaillees en moyenne par semaine dans l emploi principal, decl
"heuretra_calculees"	nombre d heures travaillees en moyenne par semaine, calcul nb jour x 7
"diff_heuretra"	difference heuretra - heuretra_calculees
"heuretra_tranche"	nombre d heures travaillees en moyenne par semaine dans l emploi principal, en t
"typehoraire"	type d horaires effectues dans l emploi principal
"heuretrasouhait"	nombre heure de travail souhaite dans l ideal, avec variation associee de revenu
"compenheuresup"	nature de compensation auxquelles donnent droit les heures supplementaires
"jourtra"	nombre de jours habituellement travailles par semaine

We exclude heuretra, jourtra and salmensuel since salhoraire is already built on wage/time worked, creating a model would lose part of its interest. However, we can try to include typehoraire and compenheuressup to control for job-specific characteristics and compensation schemesãsuch that naturally influence the hourly wage. (PHRASE POURRIE TROUVER MEILLEURS ARGUMENTS)

Des formes fonctionnelles particuliû´res:

It is likely that age and experience have non-linear relationships with hourly wages, as their marginal impact typically increases during the early career before eventually diminishing. We will test whether a quadratic specification is appropriate to capture these diminishing returns.

```{r}
#courbe 
library(ggplot2)
ggplot(df_q1, aes(age, logsalhoraire)) + 
  geom_smooth(method = "lm", formula = y ~ poly(x, 2))#???revoir
```
```{r}
plot(df_q1$age, df_q1$salhoraire,
     xlab = "Age", ylab = "Hourly wage",
     main = "Relation between Age and Hourly Wage",
     pch = 19, col = "blue", cex = 0.5)
```
```{r}
plot(df_q1$age, df_q1$logsalhoraire,
     xlab = "Age", ylab = "log of Hourly wage",
     main = "Relation between Age and log of Hourly Wage",
     pch = 19, col = "blue", cex = 0.5)
```

The concave, bell-shaped pattern of the scatter plot  suggests testing a quadratic specification for age in the Level model and the Log Level one.

```{r}
summary(lm(salhoraire ~ age + I(age^2), data=df_q1))
```
Le coefficient associûˋ û  age^2 n'est pas significatif û  aucun seuil ----- donc quadratique û  nuancer on ne sais pas ??
```{r}
plot(df_q1$expe_specifique, df$salhoraire,
     xlab = "experience", ylab = "Hourly wage",
     main = "Relation between Experience and Hourly Wage",
     pch = 19, col = "blue", cex = 0.5)
```
```{r}
plot(df_q1$expe_specifique, df$logsalhoraire,
     xlab = "experience", ylab = "Log Hourly wage",
     main = "Relation between Experience and log Hourly Wage",
     pch = 19, col = "blue", cex = 0.5)

```

The quadratic form makes less sens for experience in the level and log level models.


```{r}
names(df_q1)
```


Tests de rûˋgression:
```{r}
# 1. Level-Level Model (Baseline)

reg1_level <- lm(salhoraire ~ origine * immi + age + I(age^2) + homme + 
                as.factor(education) + expe_specifique + 
                as.factor(matri) + as.factor(csp_pere) + as.factor(typehoraire) + 
                zus + as.factor(tuu_r), data = df_q1)
summary(reg1_level)

# 2.avec en + dûˋpartement potentiellement redondant avec tuu_r
reg2_level <- lm(salhoraire ~ origine * immi + age + I(age^2) + homme + 
                as.factor(education) + expe_specifique + 
                as.factor(matri) + as.factor(csp_pere) + as.factor(typehoraire) + 
                zus + as.factor(tuu_r) + dep, data = df_q1)

summary(reg2_level)
# 3.avec en + rûˋgion au lieu de dûˋpartement
reg3_level <- lm(salhoraire ~ origine * immi + age + I(age^2) + homme + 
                as.factor(education) + expe_specifique + I(expe_specifique^2) + 
                as.factor(matri) + as.factor(csp_pere) + as.factor(typehoraire) + 
                zus + as.factor(tuu_r) + region, data = df_q1)
summary(reg3_level)

# 4.avec expe en quadratique (par intuition ûˋconomique malgrûˋ forme du nuage de point ci-avant)
reg4_level <- lm(salhoraire ~ origine * immi + age + I(age^2) + homme + 
                as.factor(education) + expe_specifique + I(expe_specifique^2) + 
                as.factor(matri) + as.factor(csp_pere) + as.factor(typehoraire) + 
                zus + as.factor(tuu_r), data = df_q1)
summary(reg4_level)
#ûÏa n'amûˋliore pas vraiment ni R**2 ni significativitûˋ statistique ou mûˆme ûˋconomique

# 5. avec taille_menage
reg5_level <- lm(salhoraire ~ origine * immi + age + I(age^2) + homme + 
                as.factor(education) + expe_specifique + I(expe_specifique^2) + 
                as.factor(matri) + as.factor(csp_pere) + as.factor(typehoraire) + 
                zus + as.factor(tuu_r) + taille_menage, data = df_q1)

summary(reg5_level)
# c'est mieux on le garde taille_menage


```






```{r}
# 2. Log-Level Model (core model)
reg1_log <- lm(log(salhoraire) ~ origine * immi + age + I(age^2) + homme + 
              as.factor(education) + expe_specifique + 
              as.factor(matri) + as.factor(csp_pere) + as.factor(typehoraire) + 
              zus + as.factor(tuu_r), data = df_q1)

# Compare the two using AIC (Lower is better)
# Note: To compare AIC of log vs level, you must adjust the log-AIC manually,
# but usually, we choose based on the distribution of residuals.?????? revoir
summary(reg1_log)
```
```{r}
models <- list(
  "reg1_log" = reg1_log,
  "reg1_level" = reg1_level,
  "reg2_level" = reg2_level
)

modelsummary(models, 
             stars = TRUE, 
             title = "Rûˋsultats des rûˋgressions")
```


```{r}
mod <- lm(logsalhoraire ~ origine + tuu, data = df_q1)
plot(mod, which = 1)
```

