---
title: "Project Econometrics - essais Aurélian 1"
author: "Aurélian"
date: "2026-04-08"
output: html_document
---

We install the general packages we need for the data manipulations
```{r}
library(haven) # to open a .dta file
library(tidyverse)
```

We import the data
```{r}
df <- read_dta("DM_Subject_2_Data.dta")
```

We recreate, correctly, the "salhoraire" variable
```{r}
data$salhoraire = data$salmensuel / (data$heuretra * 4.33)
```

Question 1

In the general case, the wage is registered for the first and the last quarters (T=1,2),
among six quarters (t=1,2,3,4,5,6).
Ici, we restrict us to the last quarter (ie, t=6).

We use filter to apply this restriction to our dataframe.
```{r}
df_q1 <- df %>% 
  filter(t==6)
```

1a) Estimation of the effect of some variables on wages

```{r}
summary(df_q1)

df_q1 <- df_q1 %>% 
  mutate(nat = as.numeric(nat),
         origine = as.numeric(origine),
         csp_actif = as.numeric(csp_actif))

reg1 <- lm(salhoraire ~ nat + origine + age + homme + csp_actif, df_q1)
summary(reg1)


```
